---
title: "Simulation: Variance Reduction Techniques"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Simulation: Variance Reduction Techniques

All these four are unbiased.

## 1.1.Antithetic Variables

**Idea**: For each Normal variate $Z_i$, consider its negative $-Z_i$:

-   Both samples come from the same distribution but are dependent

-   More generally, for Uniform(0,1) use $U_i$ and $1-U_i$

-   Calculate discounted payoff, denoted by $Y$, under both paths:
    $$ Y_i = f(Z_i), \quad \tilde{Y}_i = f(-Z_i)$$

-   Estimate price as:
    $$\overline{Y}_{AV} = \frac{1}{2n}\left(\sum_{i=1}^n Y_i + \sum_{i=1}^n \tilde{Y}_i\right) = \frac{1}{n}\sum_{i=1}^n \frac{Y_i + \tilde{Y}_i}{2}$$

    The method balances payoffs of paths with "opposite" returns.

```{r}

# Example for simple European Call

TT=1 # maturity
S0=K=50 # strike price
v=.2 # volatility
r=.03 # risk free rate

N=c(100, 500, 1000, 5000, 10000, 50000, 100000)/2

MC.mean=matrix(0,7,2) # MC estimate
MC.stde=matrix(0,7,2) # MC standard error

for(i in 1:length(N)){
  n=N[i]
  # Simple MC
  Z=rnorm(2*n)
  ST=S0*exp( (r-v^2/2)*TT + v*sqrt(TT)*Z )
  call_payoff=exp(-r*TT)*pmax(ST-K,0);put_payoff=exp(-r*TT)*pmax(K-ST,0)
  payoff = call_payoff
  MC.mean[i,1]=mean(payoff) 
  MC.stde[i,1]=sd(payoff)/sqrt(2*n) 
  
  # Antithetic Variables MC
  Z=rnorm(n); Za=-Z
  ST=S0*exp( (r-v^2/2)*TT + v*sqrt(TT)*Z )
  STa=S0*exp( (r-v^2/2)*TT + v*sqrt(TT)*Za )
  call_payoff=( exp(-r*TT)*pmax(ST-K,0) + exp(-r*TT)*pmax(STa-K,0) )/2
  put_payoff = ( exp(-r*TT)*pmax(K-ST,0) + exp(-r*TT)*pmax(K-STa,0) )/2
  payoff = call_payoff
  MC.mean[i,2]=mean(payoff) 
  MC.stde[i,2]=sd(payoff)/sqrt(n) 
}
MC.mean
MC.stde


# Compare with BSM price 
BSprice=function(S, X, r, M, v){
  # arguments: Asset price (S), Strike (X), risk-free rate (r)
  # Maturity (M), volatility (v)
  
  d1 <- (log(S/X)+(r+0.5*v^2)*M)/(v*sqrt(M))
  d2 <- d1-v*sqrt(M)
  
  call.pr= S*pnorm(d1) - X*exp(-r*M)*pnorm(d2)
  put.pr=X*exp(-r*M) * pnorm(-d2) - S*pnorm(-d1)
  
  return(cbind(call.pr,put.pr)) 
}
BSprice(S0, K, r, TT, v)

```

**Variance - Antithetic Variables**

-   Simple & general technique, but not always beneficial

-   Only helps if original & antithetic variates are **negatively
    related**

The variance comparison shows:

$$\begin{align*}
\text{Var}\left[\overline{Y}_{AV}\right] &< \text{Var}\left[\frac{1}{2n}\sum_{i=1}^{2n} Y_i\right] = \frac{1}{2n}\text{Var}\left(Y_i\right) = \frac{1}{2n}\text{Var}\left[f(Z)\right] \\
\text{where} \quad \text{Var}\left[\overline{Y}_{AV}\right] &= \frac{1}{2n}\left(\text{Var}\left[Y\right] + \text{Cov}\left(Y_i, \tilde{Y}_i\right)\right)
\end{align*}$$

This variance is less than or equal to
$\frac{1}{2n}\text{Var}\left[Y_i\right]$ only if: $$\text{Cov}\left(Y_i,
\tilde{Y}_i\right) \leq 0 $$

-   The antithetic method provides variance reduction when $f(Z)$ is
    monotonic; For non-monotonic functions, the covariance may be
    positive, making the method ineffective

## 1.2.Stratification

-   **Concept**: Split random variable domain into equi-probable strata
    & draw equal number of variates from within each one
    -   Example (2 strata): draw equal number of independent $+ve$ and
        $-ve$ $Z_i$
-   Stratification ensures equal representation of each stratum in the
    RV's domain:
    -   **Always reduces variance** (though sometimes marginally)
    -   Works best when payoff function $f(Z)$ varies significantly
        across the domain
    -   Computationally challenging for multidimensional RVs

**Stratum Definition:**

Consider $m$ equi-probable Normal strata {$A_i$}:
$P(Z \in A_j) = \frac{1}{m} \quad \text{for} \quad Z \sim N(0,1), \quad j = 1,\ldots,m$

**Stratified Estimator of** $Y = f(Z)$**:**

$$\bar{Y}_{Str} = \frac{1}{m} \sum_{j=1}^m \bar{Y}^{(j)} \text{ where the stratum estimators are: } \bar{Y}^{(j)} = \frac{1}{n} \sum_{i=1}^n f(Z_i^{(j)}) $$
$$ \text{ with stratified samples: } Z_i^{(j)} \sim^{\text{iid}} N(0,1|Z_i^{(j)} \in A_j), \quad j = 1,\ldots,m$$

-   $\bar{Y}^{(j)}$ estimates the mean within stratum $j$
-   Final estimator averages across all strata means

```{r}
# Example for simple European Call

TT=1 # maturity
S0=K=50 # strike price
v=.2 # volatility
r=.03 # risk free rate
n=10000 # <<--------||

MC.mean=0 # MC estimate
MC.stde=0 # MC standard error

# Stratification MC for various # of strata
M=c(1,10,20,50,100,200,500,1000) # <<--------||
for(i in 1:length(M)){
  m=M[i]; nj=n/m
  Z=matrix(0,nj,m)
  for(j in 1:m){
    temp=runif(nj,(j-1)/m,j/m)
    Z[,j]=qnorm(temp)
  } 
  ST=S0*exp( (r-v^2/2)*TT + v*sqrt(TT)*Z )
  payoff=exp(-r*TT)* apply(ST-K,2,pmax,0) # <<--------||
  MC.mean[i]=mean( colMeans(payoff) )
  MC.stde[i]=sqrt( sum( apply(payoff,2,var)/nj ) )/m 
}
MC.mean
MC.stde
```

-   The stratified estimator $\overline{Y}_{Str}$ is an unbiased
    estimator of $E_m[f(Z)]$.
-   The stratified estimator always has lower or equal variance compared
    to the simple Monte Carlo estimator: $$Var[\overline{Y}_{Str}]
    \leq Var[\overline{Y}] \space \space where \space \space \overline{Y} = \frac{1}{nm} \sum_{i=1}^{nm} f(Z_i)$$

## 1.3.Control Variates

To estimate $E[Y] = E[f(Z)]$ using Monte Carlo: Generate iid
$Z_i \sim N(0,1)$ and use
$\overline{Y} = \frac{1}{n}\sum_{i=1}^n Y_i = \frac{1}{n}\sum_{i=1}^n f(Z_i)$
where $f(\cdot)$ represents the option's discounted payoff.

Assume we have another option with payoff $g(\cdot)$ Known price
$E[X] = E[g(Z)]$.

Idea: Use the same random variates to estimate both $E[Y]$ and $E[X]$,
but adjust $\overline{Y}$ to take into account the error of estimate
$\overline{X}$. e.g. if $\overline{X}$ underestimate $E[X]$, then
increase $\overline{Y}$

-   Adjusted $\overline{Y}$ for estimation error $\overline{X} - E[X]$
    linearly, as
    $\overline{Y}(b) = \overline{Y} - b\left(\overline{X} - E[X]\right)$,
    where $b$ is the control coefficient

-   $\overline{Y}(b)$ is unbiased for any $b$ (provided
    $\overline{Y}, \overline{X}$ are unbiased)

-   $\mathbb{V}(\overline{Y}(b)) = \frac{1}{n}\left[\sigma_Y^2 + b^2\sigma_X^2 - 2b\sigma_{XY}\right]$

-   Optimal Coefficient : $b^* = \frac{Cov[X, Y]}{Var[X]}$

-   In practice, when $Cov[X,Y]$ and $Var[X]$ are unknown, we estimate
    b\* using MC sample:
    $\hat{b} = \frac{\sum_{i=1}^n (X_i - \overline{X})(Y_i - \overline{Y})}{\sum_{i=1}^n (X_i - \overline{X})^2}$

-   The optimal variance is:
    $Var[\overline{Y}(b^*)] = Var[\overline{Y}](1 - \rho_{XY}^2)$, where
    $\rho_{XY}$ is the correlation between $X$ and $Y$. Good control
    variates have high absolute correlation with option payoff (high
    $|\rho_{XY}|$)

```{r}
BSprice=function(S, X, r, M, v){ # Black-Scholes model
  d1 <- (log(S/X)+(r+0.5*v^2)*M)/(v*sqrt(M))
  d2 <- d1-v*sqrt(M)
  call.pr= S*pnorm(d1) - X*exp(-r*M)*pnorm(d2)
  put.pr=X*exp(-r*M) * pnorm(-d2) - S*pnorm(-d1)
  return(cbind(call.pr,put.pr)) 
}
(exact_price=BSprice(S0,K,r,M,v)[1]) 
```

MC & Control Variates

```{r}

# European call, with final asset price as control

n = 10000         # Number of simulations
r = 0.05          # Risk-free rate
v = 0.2           # Volatility
S0 = 100          # Initial asset price
K = 100           # Strike price
M = 0.5           # Time to maturity in years (6 months)

Z=rnorm(n) 

X=S0*exp( (r-v^2/2)*M + Z*v*sqrt(M) ) # control variate values
EX=S0*exp(r*M) # control variate mean

Y=pmax(X-K,0)*exp(-r*M) # payoff values (euro call)
rho_XY=cor(X,Y) # correlation between X and Y
b=sd(Y)/sd(X)*rho_XY # optimal b coefficient

Ybar=mean(Y) # Simple MC estimate
Ybar_cv=Ybar-b*(mean(X)-EX) # Control variate estimate

Ysde=sd(Y)/sqrt(n) # Simple MC std error
Ysde_cv=Ysde*sqrt((1-rho_XY^2)) # Control variate std error

print(paste("Simple MC | estimate:",Ybar, "| std error",Ysde))
print(paste("Control variate | estimate:", Ybar_cv,"| std error:",Ysde_cv))
```

```{r}

plot(X, Y, pch=16, cex=.6, col=1,ylim=c(-10,70), main="Control Variate for European Call Option",xlab="Final Asset Price (X)", ylab="Discounted Payoff (Y)")
abline(lm(Y~X), col=2, lwd=2)
abline(v=EX, h=Ybar_cv, col=3,lty=2)
legend("topleft",legend=c("Simulated Payoffs", "Regression Line", "EX & Ybar_cv Lines"),col=c(1,2, 3),pch=c(16, NA, NA),lty=c(NA, 1, 2),lwd=c(NA, 2, 1),pt.cex=c(0.6, NA, NA),bty="n")
```

## 1.4.Importance Sampling

Idea: attempt to reduce variance by **changing the distribution
(probability measure)** from which paths are generated. Change measure
to give more weight to important outcomes, thereby increasing sample
efficiency. E.g. for European call, put more weight to paths with
positive payoff (i.e. for which we exercise). Performance of importance
sampling relies heavily on equivalent measure being used.

phi: $\varphi$ \| psi: $\psi$

We want to estimate:
$\alpha = E_\varphi [f(Z)] = \int_z f(z)\varphi(z)dz$, where
$\varphi(z)$ is the probability density function (pdf) of $Z$ (Normal in
our case).

-   Simple MC: Generate i.i.d. samples from $\varphi$: and compute the
    simple estimator: $\hat{\alpha} = \frac{1}{n}\sum_{i=1}^n f(Z_i)$

-   Assume you have sample $Z_i' \sim^{iid} \psi, \quad i=1,\ldots,n$
    from new pdf $\psi$, can still estimate $\alpha$ as follows:
    $\alpha = \int_z f(z)\varphi(z)dz = \int_z f(z) \frac{\varphi(z)}{\psi(z)} \psi(z)dz = E_\psi \left[f(Z') \frac{\varphi(Z')}{\psi(Z')}\right] => \hat{\alpha}' = \frac{1}{n} \sum_{i=1}^n f(Z_i') \frac{\varphi(Z_i')}{\psi(Z_i')}$

-   The importance sampling estimator:
    $\hat{\alpha}' = \frac{1}{n}\sum_{i=1}^{n}f(Z_i') \frac{\varphi(Z_i')}{\psi(Z_i')}$
    is unbiased for $\alpha = E_\varphi[f(Z)]$ when the simple MC
    estimator $\hat{\alpha}$ is unbiased.

    -   $E_{\psi}[\hat{\alpha}'] = E_{\psi}\left[\frac{1}{n}\sum_{i=1}^{n}f(Z_i') \frac{\varphi(Z_i')}{\psi(Z_i')}\right] = \frac{1}{n}\sum_{i=1}^{n}E_{\psi}\left[f(Z_i') \frac{\varphi(Z_i')}{\psi(Z_i')}\right]= \int_{-\infty}^{\infty}f(z')\frac{\varphi(z')}{\psi(z')}\psi(z')dz'$

        $= \int_{-\infty}^{\infty}f(z')\varphi(z')dz' = E_{\varphi}[f(Z)] = \alpha$

-   The variance of $\hat{\alpha}'$ is:
    $Var_{\psi}[\hat{\alpha}'] = Var_{\psi}\left[\frac{1}{n}\sum_{i=1}^{n}f(Z_i') \frac{\varphi(Z_i')}{\psi(Z_i')}\right] = \frac{1}{n}Var_{\psi}\left[f(Z') \frac{\varphi(Z')}{\psi(Z')}\right]$

    $= \frac{1}{n}\left\{E_{\psi}\left[f^2(Z')\frac{\varphi^2(Z')}{\psi^2(Z')}\right] - \alpha^2\right\}$

-   The importance sampling estimator $\hat{\alpha}'$ has lower variance
    than the standard Monte Carlo estimator $\hat{\alpha}$ if and only
    if:
    $E_{\varphi} \left[ f^2(Z) \frac{\varphi(Z)}{\psi(Z)} \right] \leq E_{\varphi} \left[ f^2(Z) \right]$

-   $Var_{\psi}[\hat{\alpha}'] = 0 \quad \text{if} \quad \psi(z) \propto f(z)\varphi(z), \text{ for positive } f(z)$
    =\> Importance sampling works best when the new pdf $\psi$
    "resembles" the product of payoff and original density
    $f \times \varphi$.

**Code Example**

-   Consider deep out-of-the-money European call; e.g. S0 = 65, K=65.
    With simple MC, generate final prices as
    $S_T = S_0 e^{Z}, \quad \text{where } Z \sim \varphi = N\left(\left(r - \frac{\sigma^2}{2}\right)T, \sigma^2 T\right)$

-   What would be a good candidate for $\psi$?

    $\psi_1 = N\left(\log\left(\frac{90}{50}\right) - \frac{\sigma^2}{2}T, \sigma^2 T\right)$
    or
    $\psi_2 = N\left(\log\left(\frac{30}{50}\right) - \frac{\sigma^2}{2}T, \sigma^2 T\right)$

    For the first, the option will likely end up in the money.

Here, because the option is deep out-of-the-money, the probability of a
payoff is small, making vanilla Monte Carlo inefficient (rare payoff
events =\>high variance). Importance Sampling addresses this by biasing
the sampling toward more “relevant” outcomes.

```{r}
##### 
# Importance Sampling

n = 10000         # Number of simulations
r = 0.02          # Risk-free rate
v = 0.2           # Volatility
S0 = 50           # Initial asset price
K = 65            # Strike price (note: quite a bit above S0 — deep OTM)
M = 1             # Time to maturity in years (6 months)


## Plots log return Z. where Z~Nomral
z=seq(-.5,1.5,by=.001)

# Plot density under risk-neutral measure (distr. )
plot(z, dnorm(z, mean=(r - v^2/2) * M, sd=v * sqrt(M)), type='l', lwd=2)

# Scaled payoff
lines(z, pmax(S0 * exp(z) - K, 0)/S0, col=2, lwd=2)

# Product of density and scaled payoff (phi * payoff)
lines(z, dnorm(z, mean=(r - v^2/2) * M, sd=v * sqrt(M)) * pmax(S0 * exp(z) - K, 0)/2, col=4, lwd=2)

# Density under importance sampling measure (psi distr.)
lines(z, dnorm(z, mean=log(90/S0) - v^2/2 * M, sd=v * sqrt(M)), col=3, lwd=2)
lines(z, dnorm(z, mean=log(30/S0) - v^2/2 * M, sd=v * sqrt(M)), col=5, lwd=2)


legend("topright", legend = c("phi(Z)", "Scaled payoff f(Z)",
"phi(Z)*f(Z)", "psi1(Z) with 90","psi2(Z) with 30"), col = c(1, 2, 4, 3,5), lwd = 2, cex = 0.8)
```

```{r}

# similarly to previous plot, but in terms of final asset price. ST ~ LogNormal

s=seq(30,120,by=.1) 

# Plot density under risk-neutral measure (phi dist. )
plot( s, dlnorm(s/S0,meanlog=(r-v^2/2)*M,sdlog=v*sqrt(M))*13, 
      xlab="", ylab="", type='l', lwd=2)

# Scaled payoff
lines(s, pmax(s - K, 0), col = 2, lwd = 2)

# Product of density and payoff (phi * payoff)
lines( s, dlnorm(s/S0,meanlog=(r-v^2/2)*M,sdlog=v*sqrt(M))*pmax(s-K,0)*10, lwd=2, col=4 ) 

# Importance sampling density (psi distr.)
lines( s, dlnorm(s/S0,meanlog=log(90/S0)-v^2/2*M,sdlog=v*sqrt(M))*13, 
       xlab="", ylab="", type='l', lwd=2, col=3)
lines( s, dlnorm(s/S0,meanlog=log(30/S0)-v^2/2*M,sdlog=v*sqrt(M))*13, 
       xlab="", ylab="", type='l', lwd=2, col=5)

legend("topright", legend = c("phi(ST) (Lognormal)","Call payoff - f(ST) = (ST-K)+", "phi(ST)*f(ST)","psi1(ST)with90","psi2(ST)with30"),col = c(1, 2, 4, 3,5), lwd = 2, cex = 0.8)

```

```{r}
# Simulation 

mu.phi=(r-v^2/2)*M; sig.phi=v*sqrt(M)
mu.psi=log(90/50)-v^2/2*M; sig.psi=sig.phi 
Z=rnorm(n)

# Simple MC (sample from the true (phi dist.) and compute discounted payoff)
Z.phi=mu.phi+Z*sig.phi # log-returns
S.phi=S0*exp(Z.phi) # asset prices
payoff.phi=pmax(S.phi-K,0)*exp(-r*M)
MC.mean=mean(payoff.phi)
MC.sde=sd(payoff.phi)/sqrt(n)

# Importance Sampling
Z.psi=mu.psi+Z*sig.psi # generate Z under psi measure
S.psi=S0*exp(Z.psi) # final prices under psi measure
mean(S.psi) # confirm mean of final prices under psi is ~90
payoff.psi=pmax(S.psi-K,0)*exp(-r*M)
ratio=dnorm(Z.psi,mu.phi,sig.phi)/dnorm(Z.psi,mu.psi,sig.psi) # pdf ratio
IS.mean=mean(payoff.psi*ratio)
IS.sde=sd(payoff.psi*ratio)/sqrt(n)


print(paste("Simple MC | estimate:",MC.mean, "| std error",MC.sde))
print(paste("Importance Sampling | estimate:", IS.mean,"| std error:",IS.sde))
```

Recall, the exact price is:

```{r}
(exact_price=BSprice(S0,K,r,M,v)[1])
```

# 2.Gap Call Option Monte Carlo Simulation

A European gap option has a strike price K1 and a trigger price K2. The
trigger price determines whether or not the gap option is exercised,
while the strike price determines the amount of the payoff. For example,
the European gap call option has payoff given by:

$\text{Payoff} =
\begin{cases}
S_T - K_1, & \text{if } S_T \geq K_2 \\
0, & \text{if } S_T < K_2
\end{cases}$ ; Obviously, if the strike price is equal to the trigger
price then the gap option is an ordinary option. Note that gap options
can result in negative payoff at expiration (so they should not really
be called “options”). Moreover, gap options are different from barrier
options, in that for barrier options the barrier can be reached prior to
expiration, whereas for gap options only the final stock price is
compared to the trigger.

The stock price follows:
$$ S_T = S_0 e^{Z}, \quad Z \sim N\left(\left(r - \frac{\sigma^2}{2}\right)T, \sigma^2 T\right) $$Discount
factor: $e^{-rT}$

## 2.1.Basic Monte Carlo Estimation

Estimate the price of a European gap call option with: - $S_0 = 55$ -
$K_1 = 50$ - $K_2 = 60$ - $T = 1 \text{ yr}$ - $r = 2\%$ -
$\sigma = 20\%$. Assuming the price follows Geometric Brownian Motion
(GBM). Use $n = 100,\!000$ paths and calculate the standard error of
your result. Compare that to the exact price of 6.129964.

```{r}

n=100000; S0=55; K1=50; K2=60; M=1; r=.02; v=.2

Z=rnorm(n)
ST=S0*exp( (r-v^2/2)*M + v*sqrt(M)*Z)
payoff_2a=exp(-r*M)*(ST-K1)*(ST>K2)

(mean_2a=mean(payoff_2a))
(se_2a=sd(payoff_2a)/sqrt(n))

```

## 2.2.Antithetic Variables

Repeat using antithetic variables, and report: New price estimate and
Standard error

```{r}
n=100000; S0=55; K1=50; K2=60; M=1; r=.02; v=.2

Z=rnorm(n/2); Z=cbind(Z,-Z)
ST=S0*exp( (r-v^2/2)*M + v*sqrt(M)*Z)
payoff_2b=rowMeans(exp(-r*M)*(ST-K1)*(ST>K2))
(mean_2b=mean(payoff_2b))
(se_2b=sd(payoff_2b)/sqrt(n))
```

## 2.3.Stratified Sampling

Repeat using stratification with: $m = 5,10, 20$ equiprobable strata.
Report for each case: New price estimate and Standard error

```{r, warning=FALSE}

n=100000; S0=55; K1=50; K2=60; M=1; r=.02; v=.2

m.all=c(5,10,20) #<<------||
mean_2c=se_2c=rep(0,length(m.all))
for(i in 1:length(m.all)){
  m=m.all[i]; nj=n/m
  Z=matrix(0,nj,m)
  for(j in 1:m){
    temp=runif(nj,(j-1)/m,j/m)
    Z[,j]=qnorm(temp)
  } 
  ST=S0*exp( (r-v^2/2)*M + v*sqrt(M)*Z ) #<<------||
  payoff_2c=exp(-r*M)*(ST-K1)*(ST>K2) #<<------||
  mean_2c[i]=mean( colMeans(payoff_2c) )
  se_2c[i]=sqrt( sum( apply(payoff_2c,2,var)/nj ) )/m 
}
cbind( mean_2c, se_2c)
```

# 3.Forward Start European Call Option

A forward start option is an option that starts at a specified future
date with an expiration date set even further in the future.

-   For example, consider a forward start call option with start date in
    1yr and expiration date in 2yrs:

    -   at t = 0 you pay the option premium, and at t = 1 you receive a
        European call option with expiration date t = 2 and strike price
        K = S1 (this is called an at-the-money forward start call,
        because the strike price is set to be the asset price at t = 1,
        which is unknown at t = 0).

## 3.1.Basic Monte Carlo Estimation

Estimate the price of an at-the-money forward start European call option
with $T_{\text{exp}} = 2\text{yr}$, $r = 2\%$, $\sigma = 20\%$ and
assuming the price follows GBM. Use $n = 10,000$ paths and calculate the
standard error of your result. You should use the Black- Scholes formula
to explicitly calculate the value of the option at $T_{\text{start}}=1$
, i.e. you should generate $S_1$ and then calculate its “payoff” as the
Black-Scholes price of a European at-the-money call option with 1 year
to maturity. Compare that to the exact price of 6.241225.

```{r}

n=10000; S0=70; Tstart=1; Texp=2; r=.02; v=.2

Z=rnorm(n)
STstart=S0*exp( (r-v^2/2)*Tstart + v*sqrt(Tstart)*Z )
payoff_3a = exp(-r*Tstart) * BSprice(STstart, STstart, r, Texp-Tstart, v)[,1]
(mean_3a=mean(payoff_3a)) # MC estimate
(se_3a=sd(payoff_3a)/sqrt(n))
```

## 3.2.Antithetic Variables

Repeat using antithetic variables, and report the new price estimate and
standard error

```{r}
n=10000; S0=70; Tstart=1; Texp=2; r=.02; v=.2

Z=rnorm(n/2); Z.a=-Z
STstart=S0*exp( (r-v^2/2)*Tstart + v*sqrt(Tstart)*Z )
STstart.a=S0*exp( (r-v^2/2)*Tstart + v*sqrt(Tstart)*Z.a ) # antithetic path
payoff=exp(-r*Tstart) * BSprice(STstart, STstart, r, Texp-Tstart, v)[,1]
payoff.a=exp(-r*Tstart) * BSprice(STstart.a, STstart.a, r, Texp-Tstart, v)[,1] # antithetic path payoff
payoff_3b=(payoff+payoff.a)/2
(mean_3b=mean(payoff_3b)) # MC estimate
(se_3b=sd(payoff_3b)/sqrt(length(payoff_3b))) # std. error
```

## 3.3.Stratification Sampling

Repeat part using stratification with equiprobable strata, and report
the new price estimate and standard error.

```{r}
n=10000; S0=70; Tstart=1; Texp=2; r=.02; v=.2

m=10; nj=n/m  # Samples per stratum
Z=STstart=payoff_3c=matrix(0,nj,m)
for(j in 1:m){
  temp=runif(nj,(j-1)/m,j/m) # Uniforms in [ (j-1)/m, j/m ]
  Z[,j]=qnorm(temp) # Map to standard normal via inverse CDF
  STstart[,j]=S0*exp( (r-v^2/2)*Tstart + v*sqrt(Tstart)*Z[,j] ) # Asset at Tstart
  payoff_3c[,j]=exp(-r*Tstart) * BSprice(STstart[,j], STstart[,j], r, Texp-Tstart, v)[,1]
} 
(mean_3c=mean( colMeans(payoff_3c) ) )
(se_3c=sqrt( sum( apply(payoff_3c,2,var)/nj ) )/m )
```

Repeat using stratification with: $m = 5,10, 20$ equiprobable strata.
Report for each case: New price estimate and Standard error

```{r}

n=10000; S0=70; Tstart=1; Texp=2; r=.02; v=.2
m.all = c(5, 10, 20)

mean_3c = se_3c = rep(0, length(m.all))
for (i in 1:length(m.all)) {
  m = m.all[i]; nj = n / m
  Z = matrix(0, nj, m); ST = matrix(0, nj, m)
  payoff_3c = matrix(0, nj, m)
  for (j in 1:m) {
    temp = runif(nj, (j-1)/m, j/m)
    Z[,j] = qnorm(temp)
  }

  # Simulate ST at Tstart
  ST = S0 * exp((r - v^2/2) * Tstart + v * sqrt(Tstart) * Z)

  # Forward-start call payoff: BS price at Tstart, strike = ST
  for (j in 1:m) {
    payoff_3c[,j] = exp(-r * Tstart) * BSprice(ST[,j], ST[,j], r, Texp - Tstart, v)[,1]
  }

  mean_3c[i] = mean(colMeans(payoff_3c))
  se_3c[i] = sqrt(sum(apply(payoff_3c, 2, var) / nj)) / m
}

cbind(mean_3c, se_3c)

```

## 3.4.Analytical Solution

Show that the price of the at-the-money forward call option is actually
a simple linear function of $S_0$

Hint: Use risk-neutral pricing and express the payoff at
$T_{\text{start}}$ as the Black-Scholes price of an ATM call with
maturity $T > T_{\text{start}}$

```{r}
Dt=Texp-Tstart
d1=(r/v+v/2)*sqrt(Dt)
d2=(r/v-v/2)*sqrt(Dt)
(exact_price=S0*( pnorm(d1) - exp(-r*Dt)*pnorm(d2) ))
```

# 4.a-Quantile Option Pricing

An $\alpha$-quantile option is an option whose payoff is determined by
the $\alpha$th quantile of the underlying price from time 0 until
expiration. For example, the median call option (i.e., $\alpha=50\%$)
with strike $K$ has payoff
$\left(\text{median}\left\{S_t\right\}_{t=0}^T-K\right)_+$, where median
$\left\{S_t\right\}_{t=0}^T$ is the median asset price over $[0,T]$.
Assume that the underlying asset follows GBM with $S_0 = 60$, $r = 5\%$,
$\sigma = 25\%$, and use simulation to estimate the price of median call
with $K = 70$ & $T = 1$.

## 4.1.Path discretization - Simple Monte Carlo

Perform simple Monte Carlo using path discretization with $t_j = j*T/m$,
$j = 0,\ldots,m$ and $m = 50$ in order to approximate the continuous
median $\left\{S_t\right\}_{t=0}^T$ by the discrete median
$\left\{S(t_j)\right\}_{j=0}^m$. Use $n = 10,000$ paths and report:
Price estimate and Standard error of the estimate

-   `S0*exp((r-v^2/2)*M+v*sqrt(M)*Z)` : Simulates the asset price at a
    single future time M. It assumes no intermediate steps — just jump
    directly from t=0 to t=M

-   `S = S0 * exp(t(apply((r - v^2/2)*Dt + Z*v*sqrt(Dt), 1, cumsum)))`:
    Simulates the entire path of the asset price over multiple time
    steps $\Delta t =M/m$, where m is the number of steps.
    $(r - \frac{1}{2} v^2) \Delta t + v \sqrt{\Delta t} Z_t$

    -   At each time step $t_i$, simulate:

        $\log(S_{t_i}) = \log(S_{t_{i-1}}) + \left( r - \frac{1}{2} v^2 \right) \Delta t + v \sqrt{\Delta t} Z_i$

        Taking the cumulative sum and exponentiating gives the entire
        path:

        $S_t = S_0 \cdot \exp \left( \sum_{i=1}^t \left( (r - \frac{1}{2} v^2) \Delta t + v \sqrt{\Delta t} Z_i \right) \right)$

-   `quantile()` function and specify the desired probability level.

```{r}

S0 = 60       # Initial stock price
K = 70        # Strike price
M = 1         # Time to maturity
r = 0.05      # Risk-free rate
v = 0.25      # Volatility

n = 10000     # Number of simulated paths
m = 50        # Number of time steps per path
Dt = M/m      # Time step size

Z=matrix(rnorm(n*m), n, m) # simulate n*m standard normal for Brownian increments.
S=S0*exp(t(apply((r-v^2/2)*Dt+Z*v*sqrt(Dt),1,cumsum))) # <<<--------|||
# q=0.5; quantS = apply(S, 1, quantile, probs = q)
medS=apply(S,1,median)
payoff_4a=exp(-r*M)*pmax(medS-K,0) # <<<--------|||
(mean_4a=mean(payoff_4a))
(se_4a=sd(payoff_4a)/sqrt(n))
```

## 4.2.Control Variate (BS price of European Call)

Repeat using the Black-Scholes price of the European call as a control
variate, and report the new price estimate and standard error

```{r}
EX=BSprice(S0, K, r, M, v)[1] 
X=exp(-r*M)*pmax(S[,m]-K,0) # control variate values
Y=payoff_4a

(rho_XY=cor(X,Y)) # correlation between X and Y
(b=sd(Y)/sd(X)*rho_XY) # optimal b coefficient
(mean_4b=mean_4a - b*(mean(X)-EX)) # Control variate estimate
(se_4b=se_4a*sqrt((1-rho_XY^2))) # Control variate std error
```

-   `pmax(S[,m]-K,0)`: payoff of a European call option at maturity

```         
-   `S` is an n × m matrix representing the simulated stock price
    paths. m is number of steps.

-   `S[,m]` means: the value of each path at the final time step
```

# 5.Asian option Pricing (average)

An Asian option is an option whose payoff is determined by the
**average** underlying price over some preset period of time, typically
the time till expiration. For example, the Asian call option with strike
$K$ has payoff $(A(0,T)-K)_{+}$, where $A(0,t)$ is the **average asset
price** over $(0,T)$. There are two types of averaging: $$ \begin{cases}
\text{arithmetic average}: A(0,T)=\dfrac{1}{T}\int_{0}^{T}S(t)dt \\ 
\\
\text{geometric average}: A(0,T)=\exp\left(\dfrac{1}{T}\int_{0}^{T}\ln\left(S(t) \right)dt\right)
\end{cases} $$ If the underlying asset follows GBM, then the price of an
Asian option with geometric averaging can be calculated explicitly. For
Asian options with arithmetic averaging, however, we rely exclusively on
numerical methods for pricing.

Assume that the underlying asset follows GBM with $S_{0}=70$, $r=5\%$,
$\sigma=25\%$, and use simulation to estimate the price of an arithmetic
Asian call with $K=80$ & $T=1$.

## 5.1.path discretization - Simple Monte Carlo

Perform simple Monte Carlo using path discretization with $m = 50$ in
order to approximate\
$A(0, T) = \frac{1}{T} \int_{0}^{T} S(t) dt$ by
$\tilde{A}(0, T) = \frac{1}{m} \sum_{j=1}^{m} S(t_{i})$. Use
$n = 10,000$ paths and report: Price estimate and Standard error of the
estimate

```{r}
set.seed(12345)
S0 = 70       # Initial stock price
K = 80        # Strike price
M = 1         # Time to maturity
r = 0.05      # Risk-free rate
v = 0.25      # Volatility

n = 10000     # Number of simulated paths
m = 50        # Number of time steps per path
Dt = M/m      # Time step size

Z=matrix(rnorm(n*m), n, m) # simulate n*m standard normal for Brownian increments.
S=S0*exp(t(apply((r-v^2/2)*Dt+Z*v*sqrt(Dt),1,cumsum))) # <<<--------|||
# q=0.5; quantS = apply(S, 1, quantile, probs = q)
meanS=apply(S,1,mean) # meanS=rowMeans(S)
payoff_5a=exp(-r*M)*pmax(meanS-K,0) # <<<--------|||
(mean_5a=mean(payoff_5a))
(se_5a=sd(payoff_5a)/sqrt(n))
```

## 5.2.Control Variate Approach

Control Variate Approach Repeat using the price of the geometric Asian
call as a control variate, and report: Price estimate and Standard error
of the estimate. Note: The exact price of the geometric Asian option is
1.271969.

```{r}

EX=1.271969 # Control variate - exact Geom. Asian call price
A.geom=exp(rowMeans(log(S)))
X=exp(-r*M)*pmax(A.geom-K,0) # control variate values
Y=payoff_5a

(rho_XY=cor(X,Y)) # correlation between X and Y
(b=sd(Y)/sd(X)*rho_XY) # optimal b coefficient

(mean_5b=mean_5a - b*(mean(X)-EX)) # Control variate estimate
(se_5b=se_5a*sqrt((1-rho_XY^2))) # Control variate std error
```

## 5.3.Importance Sampling

Repeat using importance sampling: Change the probability measure so that
$E[S_T] = K$ and Experiment with at least two more measure changes of
your choice. Report for each case: New price estimate and Standard error
of the estimate.

=\> New drift $= \frac{\log(K/S_0)}{M}$;
$E[S_T] = K = S_0 \cdot e^{\text{new drift} \cdot M}$

```{r}

MEANS=c(80,90,100) # different importance sampling drift
mean_5c=se_5c=rep(0,length(MEANS))


for(i in 1:length(MEANS)){
  new.r=log(MEANS[i]/S0)/M # new drift
  ZZ=(new.r-v^2/2)*Dt+Z*v*sqrt(Dt) # simulates log-returns under new drift
  S=S0*exp(t(apply(ZZ,1,cumsum))) # matrix of simulated stock paths over time
  A=rowMeans(S) # Arithmetic average of each path (used in payoff)
  log.ratio=dnorm(ZZ,(r-v^2/2)*Dt,v*sqrt(Dt),log=TRUE)-dnorm(ZZ,(new.r-v^2/2)*Dt,v*sqrt(Dt),log=TRUE) # Likelihood ratio (Radon-Nikodym derivative)
  prod.ratio=exp( rowSums(log.ratio) ) 
  payoff=exp(-r*M)*pmax(A-K,0)*prod.ratio # payoff under importance sampling
  mean_5c[i]=mean(payoff)
  se_5c[i]=sd(payoff)/sqrt(n)
}

cbind(mean_5c, se_5c)
```

-   `dnorm(x, mean = mu, sd = sigma, log = TRUE)`: In R, the `dnorm()`
    function computes the probability density function (PDF) of the
    normal distribution. When you set `log=TRUE`, it returns the
    logarithm of the density instead of the raw density value.

    -   This equivalent to :
        `log(1 / (sqrt(2 * pi) * sigma) * exp(- (x - mu)^2 / (2 * sigma^2)))`
        which simplifies to
        `-0.5 * log(2 * pi) - log(sigma) - (x - mu)^2 / (2 * sigma^2)`
