---
title: "Optimization in Finance"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---
This is sourced from STAD70 course practice questions and sample R code taught by professor Sotos. If you have any questions/concerns/comments feel free to email me: cristal.wang111@gmail.com.

# 1.Dynamic Programming (l/n)

Use Dynamic Programming to find optimal trading strategy under
transaction costs given price evolution

```{r,warning=FALSE}

library(tseries)

St = get.hist.quote( "SHOP.TO", start = "2023-01-01", end = "2023-03-31", quote = "Close" )
plot(St, lwd=2, type = "o", pch=16, main = "Shopify Close Price")
```

```{r}

tc = .1 # cost per transaction

n = length(St)

# optimal value function
n.val = rep(0,n)  # neutral (n) 
l.val = rep( as.numeric(St[n])-tc, n ) # long (l)

# optimal strategy
n.str = rep(0,n)
l.str = rep(1,n)

# backward induction
for( i in (n-1):1 ){
  
  n2n = n.val[i+1] 
  n2l = l.val[i+1] - as.numeric(St[i]) - tc
  if( n2n >= n2l ){  # action: n -> n
    n.val[i] = n2n
    n.str[i] = 0
  }else{             # action: n -> l
    n.val[i] = n2l
    n.str[i] = 1
    n.str[(i+1):n] = l.str[(i+1):n]
  }
    
  l2n = n.val[i+1] + as.numeric(St[i]) - tc
  l2l = l.val[i+1]
  if( l2n >= l2l ){  # action: l -> n
    l.val[i] = l2n
    l.str[i] = 0
    l.str[(i+1):n] = n.str[(i+1):n]
  }else{             # action: l -> l
    l.val[i] = l2l
  }
}

# calculate cumulative optimal value 
run.val = rep(0, n)
run.val[1] = n.str[1] * ( - as.numeric(St[1]) - tc )
for(i in 2:n){
  if( n.str[i] == n.str[i-1] ){
    run.val[i] = run.val[i-1]
  }else if( n.str[i] > n.str[i-1] ){
    run.val[i] = run.val[i-1] - as.numeric(St[i]) - tc
  }else{
    run.val[i] = run.val[i-1] + as.numeric(St[i]) - tc
  }
}

run.val[n] 
n.val[1]

# create time series
n.val = zoo::zoo( n.val, order.by = time(St))
n.str = zoo::zoo( n.str, order.by = time(St))
run.val = zoo::zoo( run.val, order.by = time(St))

# plots
plot( St, lwd=2, main = "Strategy")
points( St, col = n.str+2, pch = 16)
abline( v = time(St), lwd=.1, lty=2)

plot( run.val, lwd=2 , main = "Value of strategy")
points( run.val, col = n.str+2, pch = 16)
abline( v = time(St), lwd=.1, lty=2)


```

# 2.Dynamic Programming(l/n/s)

Consider the example of trading an asset with transaction costs,
assuming you have perfect knowledge of its price. Adapt the code from
lecture 12 to find the optimal strategy and its cumulative P/L, assuming
you can also short the asset, i.e. you have to consider three states:
long, neutral, and short. Note that when you go from long to short
position, or vice-versa, you have to pay twice the transaction cost
because you sell/buy two units of asset.

Test your code on Shopify (SHOP.TO) closing daily prices from Jan 1 to
Mar 31, 2023 with a transaction cost of \$0.1/share, and plot the
strategy.

```{r}
library(tseries)

St = get.hist.quote( "SHOP.TO", start = "2023-01-01", end = "2023-03-31", quote = "Close" )
plot(St, lwd=2, type = "o", pch=16, main = "Shopify Close Price")


tc = 1 # cost per transaction

n = length(St)

# optimal value function
n.val = rep( 0, n ) # neutral (n) 
l.val = rep( 0, n ) # long (l)
s.val = rep( 0, n ) # short (s)

l.val[n] =  as.numeric(St[n])-tc
s.val[n] = -as.numeric(St[n])-tc

# optimal strategy
n.str = rep(  0, n )
l.str = rep(  1, n )
s.str = rep( -1, n )

# Backward Induction/Dynamic Programming
# We have 3 options/transitions to consider for each state:

for( i in (n-1):1 ){

  # entering at n, compare: n2n, n2l, n2s 
  n2n = n.val[i+1] 
  n2l = l.val[i+1] - as.numeric(St[i]) - tc
  n2s = s.val[i+1] + as.numeric(St[i]) - tc
  if( n2n >= max( n2l, n2s) ){          # action: n -> n
    n.val[i] = n2n
    n.str[i] = 0
  }else if( n2l >= max( n2n, n2s) ){    # action: n -> l
    n.val[i] = n2l
    n.str[i] = 1
    n.str[(i+1):n] = l.str[(i+1):n]
  }else{                                # action: n -> s
    n.val[i] = n2s
    n.str[i] = -1
    n.str[(i+1):n] = s.str[(i+1):n]
  }
  
  # entering at l, compare: l2n, l2l, l2s 
  l2n = n.val[i+1] + as.numeric(St[i]) - tc
  l2l = l.val[i+1]
  l2s = s.val[i+1] + 2*( as.numeric(St[i]) - tc )
  if( l2n >= max( l2l, l2s ) ){        # action: l -> n
    l.val[i] = l2n
    l.str[i] = 0
    l.str[(i+1):n] = n.str[(i+1):n]
  }else if( l2l >= max( l2n, l2s ) ){  # action: l -> l
    l.val[i] = l2l
  }else{
    l.val[i] = l2s                     # action: l -> s
    l.str[i] = -1
    l.str[(i+1):n] = s.str[(i+1):n]
  }

  # entering at s, compare: s2n, s2l, s2s 
  s2n = n.val[i+1] - as.numeric(St[i]) - tc
  s2l = l.val[i+1] + 2*(-as.numeric(St[i]) - tc)
  s2s = s.val[i+1]
  if( s2n >= max( s2l, s2s ) ){        # action: s -> n
    s.val[i] = s2n
    s.str[i] = 0
    s.str[(i+1):n] = n.str[(i+1):n]
  }else if( s2l >= max( s2n, s2s ) ){  # action: s -> l
    s.val[i] = s2l
    s.str[i] = 1
    s.str[(i+1):n] = n.str[(i+1):n]
  }else{
    s.val[i] = s2s                     # action: s -> s
    s.str[i] = -1
  }
  
}

# calculate cumulative optimal value 
run.val = rep(0, n)
run.val[1] = - n.str[1] * as.numeric(St[1]) - abs(n.str[1]) * tc
for(i in 2:n){
  if( n.str[i] == n.str[i-1] ){
    run.val[i] = run.val[i-1]
  }else if( n.str[i] > n.str[i-1] ){
    run.val[i] = run.val[i-1] + (n.str[i] - n.str[i-1]) * (-as.numeric(St[i]) - tc)
  }else{
    run.val[i] = run.val[i-1] + (n.str[i-1] - n.str[i]) * (as.numeric(St[i]) - tc)
  }
}

run.val[n] 
n.val[1]

# create time series
n.val = zoo::zoo( n.val, order.by = time(St))
n.str = zoo::zoo( n.str, order.by = time(St))
run.val = zoo::zoo( run.val, order.by = time(St))

# plots
tmp = 1 + n.str * 2
tmp = replace( tmp, n.str == -1, 2 )

plot( St, lwd=2, main = "Strategy")
points( St, col = tmp, pch = 16)
abline( v = time(St), lwd=.1, lty=2)
legend( "topleft", pch=16, col = c(2,1,3), c("short","neutral","long"))

plot( run.val, lwd=2 , main = "Value of strategy")
points( run.val, col = tmp, pch = 16)
abline( v = time(St), lwd=.1, lty=2)
legend( "topleft", pch=16, col = c(2,1,3), c("short","neutral","long"))


```
