---
title: " 3.1.Multivariate Return Modelling & Copulas"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
---

This is sourced from STAD70 course practice questions and sample R code taught by professor Sotos. If you have any questions/concerns/comments feel free to email me: [cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Multivariate Return Modelling

## 1.1.Data and Return

```{r}
library(tseries)

ticker.symbols= c("KLAC","JNJ","DUK","PNC","AAPL")
N.stocks = length(ticker.symbols)

start.date=as.Date("2003-08-01")  # Prices from Aug 2003 to Aug 2013
end.date=as.Date("2013-08-01")
P=get.hist.quote(instrument = ticker.symbols[1], start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T)
names(P)=ticker.symbols[1]

for (i in 2:N.stocks) {
  cat("Downloading ", i, " out of ", N.stocks , "\n")
  x=try(get.hist.quote(instrument = ticker.symbols[i], start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T))
  if( class(x)!= "try-error"){
    names(x)=ticker.symbols[i]
    P = merge(P, x, all=TRUE)
  }
}
```

```{r}
R=diff(log(P)) # Calculate log returns
head(R,3) #Input: data R with several columns (e.g., each column has returns of an asset)
```

## 1.2.Sample Covariance Estimation

Multivariate Normal Estimation

### 1.2.1.Scatterplot Matrix of Return

```{r}
pairs(R, pch=16, cex=.5) 
```

### 1.2.2.Sample Covariance and correlation matrix

```{r}
COV=cov(R) # sample covariance matrix
COR=cov2cor(COV) # sample correlation matrix (from covariance) 
```

### 1.2.3.Levelplot of correlation matrix

darker =\> higher correlation

```{r}
library(lattice)
levelplot( COR[,ncol(R):1]  ) # [,ncol(R):1]  means reverse, can remove
```

## 1.3.Robust estimation

Robust means it is not significantly influenced by outliers.

```{r}
library(MASS)
```

### 1.3.1.Robust estimator of covariance matrix

```{r}
out=cov.rob(R)  
COV.rob=out$cov
```

### 1.3.2.level plot of robust correlation matrix

```{r}
COR.rob=cov2cor(COV.rob)
levelplot( COR.rob[,ncol(R):1]  )
```

## 1.4.Multivariate Student's t estimation and simulation

-   Never remove outliers in finance ⇒ model heavy tails using *multivariate* t distribution
    -   *Multivariate* Normal scale mixture

$$R \sim t_v(\mu, \Lambda); \quad R = \mu + Z\sqrt{\nu / W},  \quad \text{where} \begin{cases} 
W \sim \chi^2(df = v) \\ 
Z \sim N(0, \Lambda), \\
\end{cases} \quad \text{where} \space E[R] = \mu, \space Cov[R] = \Lambda \cdot \frac{v}{v-2}$$

-   Marginals are t-distributed with same df =\> all assets returns have same tail index

-   Model exhibits tail dependence

```{r}
library(mnormt)
```

### 1.4.1.Multivariate t estimation (fit t, find df by MLE)

-   `cov.trob`: It stands for **covariance estimation using a t-distribution** (hence, “t-robust”). The method assumes your data comes from a multivariate t-distribution, not a normal distribution. Useful when your data has heavy tails or outliers.

-   `fit = cov.trob(R, nu = df.opt)`

    -   `fit$cov`: the estimated robust covariance matrix

    -   `fit$center`: the estimated mean vector

```{r}
library(MASS) # needed for cov.trob
library(mnormt) # needed for dmt

## range of degree of freedom
df = seq(1.5,8,by=.2)
n = length(df)

## for each df in range,fit robust covariance matrix, calculate the corresponding log-likelihood, store in loglik
loglik = rep(0,n)
for(i in 1:n){
  fit = cov.trob(R,nu=df[i])
  loglik[i] = sum(log(dmt(R,mean=fit$center,S=fit$cov,df=df[i])))
}
## take the df(nu) which maximize the log-likelihood
df.opt=df[which.max(loglik)]
fit = cov.trob(R,nu=df.opt)

fit$cov * (df.opt/(df.opt-2)) # t-distribution's covariance matrix df/(df-2)
levelplot( cov2cor(fit$cov)[,ncol(R):1] )# t-distribution's correlation matrix
```

### 1.4.2.Simulate multivariate t

Use $$\mathbf{t}=\boldsymbol{\mu}+\mathbf{Z}\sqrt{\frac{\nu}{W}}$$

```{r}
Sim_Number = 5000
Z=rmnorm(Sim_Number,rep(0,ncol(R)),fit$cov)
W=rchisq(Sim_Number,df.opt)
R.sim = matrix(fit$center,Sim_Number,ncol(R), byrow=T) + diag( sqrt(df.opt/W) ) %*% Z # simulated t dist
# simulated scatterplot
pairs(R.sim,pch=16,cex=.5) 
```

-   `matrix(fit$center, Sim_Number, ncol(R), byrow = TRUE)`: This creates a matrix with `Sim_Number` (e.g. 5000) rows, each being a copy of the mean vector `fit$center`.

    ```{r}
    print(dim(matrix(fit$center,Sim_Number,ncol(R), byrow=T)))
    print(head(matrix(fit$center,Sim_Number,ncol(R), byrow=T),3))
    ```

### 1.4.3.Compare QQ-plots of real & simulated returns

```{r}
for(i in 1:5){
  plot(density(R[,i]), lwd=2, main=ticker.symbols[i])
  lines(density(R.sim[,i]), lwd=2, col=2) 
  legend("topright", lwd=2 , col=1:2, c("actual","simulated") )
}

```

# 2.Copulas

Note: in this chichi, $\Phi,\Phi_{\mu,\Sigma}$ is cdf of $N(0,1),N(\mu,\Sigma)$ respectively. $C_\rho$ is guassian copula of $N(0,\rho)$.

## 2.1.Gaussian copula

### 2.1.1.Define a Gaussian copula by normal cdf, then plot contour and 3D Perspective Plot.

```{r,warning=FALSE}
# library
library(mvtnorm) # multivariate Normal utilities
# Input: correlation matrix
rho=matrix(data=0.75,nrow=2,ncol=2); diag(rho)=1
## equivalently to:
# rho = matrix(c(1,0.75,0.75,1),nrow=2)
```

$$C_\rho(u_1,u_2)=\Phi_{0,\rho}(\Phi^{-1}(u_1),\Phi^{-1}(u_2))=pmvnorm_{\rho}(qnorm(u_1),qnorm(u_2))$$

-   `qnorm(u_i)` : This is the inverse of the CDF of the standard normal. Transforms $u$ from uniform(0,1) space to standard normal quantiles. $$\Phi^{-1}(u_i)$$

```{r}
## define the gaussian copula = cdf of multivariate normal
GaussCopula=function(u,rho){return( pmvnorm(upper=qnorm(u),sigma=rho) )}  
# pmvnorm(CDF), qnorm(inverse cdf)
```

calculate the value of copula at each point u in [0,1]\*[0,1]

```{r}
n=30
u=seq(0,1,length=n)
GC=matrix(0,n,n)
for(i in 1:n){
  for(j in 1:n){
    GC[i,j]=GaussCopula(c(u[i],u[j]),rho)
  }
}
```

plot contour and 3D perspective plot

```{r}
contour(u,u,GC) # contour 
persp(u,u,GC, theta=-25, phi=20) # 3D perspective plot

```

### 2.1.2.Generating Variates from Gaussian copula

In this part:

Step 1. we first generate Normal data $Z=\begin{bmatrix}Z_1\\Z_2\end{bmatrix}\sim N(0,\rho)$. Then $(U_1,U_2)=(\Phi(Z_1),\Phi(Z_2))\sim C_\rho$.

Step 2. Let $F_1,F_2$ be two (marginal) cdf, then $(X_1,X_2)=(F_1^{-1}(U_1),F_2^{-1}(U_2))$ has meta-Gaussian distribution with marginal $F_{1},F_{2}$ and copula $C_\rho$.

```{r}
#### Input: correlation, sample size
rho=matrix(data=.75,nrow=2,ncol=2); diag(rho)=1
# rho = matrix(c(1,0.75,0.75,1),nrow=2) ## equivalently
n=1000 # simulate size
```

#### Step1.Gaussian copula variates

We first generate Normal data $Z=\begin{bmatrix}Z_1\\Z_2\end{bmatrix}\sim N(0,\rho)$. Then $(U_1,U_2)=(\Phi(Z_1),\Phi(Z_2))\sim C_\rho$.

```{r}
Z=rmvnorm(n,sigma=rho) # 2D Normal random variates
plot(Z, pch=16, xlab="",ylab="")

U=pnorm(Z) # Corresponding Gaussian copula variates
plot(U, pch=16, xlab="",ylab="")

```

-   `pnorm(Z)` → CDF (area under curve up to Z)

-   `dnorm(Z)` → PDF (density at Z, i.e., height of the bell curve)

#### Step2.Pick Marginals get Meta-Gaussian

Let $F_1,F_2$ be two (marginal) cdf, then $(X_1,X_2)=(F_1^{-1}(U_1),F_2^{-1}(U_2))$ has meta-Gaussian distribution with marginal $F_{1},F_{2}$ and copula $C_\rho$.

##### step2.eg1.Meta-Gaussian with both marginal $\chi^2(df=10)$

```{r}
X=qchisq(U,10) # Generating (meta-Gaussian) Chi-square variates
plot(X, pch=16, xlab="",ylab="")
```

##### step2.eg2.Meta-Gaussian with marginal exponential & normal

```{r}
X1=qexp(U[,1],4) # Generating (meta-Gaussian) Exponential & Normal variates
X2=qnorm(U[,2],.03,.04)
plot(X1,X2, pch=16, xlab="",ylab="")
```

## 2.2.Elliptical copulas (Gaussian copula and t copula)

**Gaussian copula** only depends on $\rho$.

**t copula** depends on $\rho$ and $df$.

```{r,warning=FALSE}
library(copula)
```

### 2.2.1 p2P and P2p (Correlation vector)

In `normalCopula()` and `tCopula()`, the input correlation should be a **vector** instead of correlation matrix.

#### Vector of correlations =\> correlation matrix

```{r}
rho_vector = c(.1,.2,.3)
rho_mat=p2P(rho_vector)
rho_mat
```

```{r}
p2P(0.6)
```

#### Correlation matrix =\> Vector

```{r}
rho_mat = matrix(c(1.0, 0.1, 0.2, 0.1,1.0,0.3,0.2,0.3,1.0), 3, 3)
P2p(rho_mat)
```

### 2.2.2.Normal copula (Gaussian copula use library)

```{r}
## Input: correlation vector (library need),sample size
rho_vector = 0.75
n=1000
```

#### (a) Construct normal copula

```{r}
Normal.cop = normalCopula( rho_vector, dim=2, dispstr="un") 
u = rCopula(n, Normal.cop)
plot(u, pch=16, cex=.8, xlab="", ylab="")
```

**3D plot (pdf), pdf and cdf contour plots**

```{r}
persp(Normal.cop,dCopula, zlab="", xlab="", ylab="") # plot of copula pdf
contour(Normal.cop,dCopula) # contour of copula pdf
contour(Normal.cop,pCopula) # contour of copula cdf
```

#### (b) Generate meta gaussian

##### e.g.1.Bivariate normal: normal copula and marginal N(0,1)

```{r}
Normal.Mvd = mvdc(Normal.cop, margins = c("norm", "norm"), 
              paramMargins = list(list(mean = 0, sd = 1), list(mean = 0, sd = 1)))
```

bivariate normal pdf 3D plot and contour plot

```{r}
contour(Normal.Mvd,dMvdc, xlim = c(-3, 3), ylim = c(-3, 3)) # contours of bivariate density

persp(Normal.Mvd,dMvdc, xlim = c(-3, 3), ylim = c(-3, 3), zlab="", xlab="", ylab="") # 3D plot of bivariate density
```

##### e.g.2.Exponential marginal meta-Gaussian data

Gives two exponential variables, each with their own rate, but introduce dependency between them via a Gaussian copula. Copulas let you separate marginals from dependency structure.

-   `Normal.cop`: The copula defines the **dependence structure** (correlation between variables).

-   `margins = rep("exp", 2)`: You want two **Exponential marginals**.

-   `paramMargins`: Each marginal has its own rate:

    -   First variable: Exponential with rate 2

    -   Second variable: Exponential with rate 3

```{r}
Exp.Mvd = mvdc(Normal.cop, margins =  rep("exp",2), 
                   paramMargins = list(list(rate=2), list(rate=3)))
Exp.sample=rMvdc(n = 1000, mvdc = Exp.Mvd)

```

### 2.2.3.t copula

```{r}
## Input: correlation vector,sample size,
rho_vector = 0.75
n=1000
df=2
```

#### (a) construct t copula, generating and plot sample data from copula

```{r}
t.cop = tCopula( rho_vector, df=df, dim = 2, dispstr="un") # 2D Student-t copula with df=2 & 0.75 correlation 
u = rCopula(n, t.cop)
plot(u, pch=16, cex=.8, xlab="", ylab="")
```

**copula pdf 3D plot, pdf contour plot**

```{r}
persp(t.cop,dCopula, zlab="", xlab="", ylab="")
contour(t.cop,dCopula)
```

#### (b) Generate multivariate distribution with t copula

##### e.g.1.Multivariate distribution with t copula & N(0,1) marginals, plot contour of pdf

```{r}
t.Mvd = mvdc(t.cop,margins = c("norm", "norm"), 
             paramMargins = list(list(mean = 0, sd = 1), list(mean = 0, sd = 1)))
contour(t.Mvd,dMvdc, xlim = c(-3, 3), ylim = c(-3, 3))

```

## 2.3.Archimedean copulas

### 2.3.1.Clayton/Frank/Gumbel copula

```{r}
## Input: copula,sample size
n=1000

## Clayton
cop = claytonCopula(param=2.5, dim=2)
## or Frank
# cop = frankCopula(param=9, dim=2)
## or Gumbel
# cop = gumbelCopula(param=2.5, dim=2)
```

generating data from copula and plot

```{r}
u = rCopula(n, cop)
plot(u, pch=16, cex=.8, xlab="", ylab="")
```

pdf 3D perspective plot, pdf and cdf contour plots

```{r}
persp(cop,dCopula, zlab="", xlab="", ylab="")
contour(cop,dCopula, nlevels=100)
contour(cop,pCopula)
```

### 2.3.2.Generate multivariate distribution with chosen (Clayton/Frank/Gumbel) copula & N(0,1) marginals, plot contour of pdf

```{r}
Mvd = mvdc(cop, margins = c("norm", "norm"), 
            paramMargins = list(list(mean = 0, sd = 1), list(mean = 0, sd = 1)))
contour(Mvd, dMvdc, xlim = c(-3, 3), ylim = c(-3, 3))
```
