---
title: "4.Portfolio Theory"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Efficient Equity Portfolios

This section uses daily stock prices in the data set `Stock_Bond.csv`
that is posted on the book’s website.

Variable name ends with “AC” is an adjusted closing price. As the name
suggests, these prices have been adjusted for dividends and stock
splits, so that returns can be calculated without further adjustments.

-   Run the following code, which will read the data, compute the
    returns for six stocks, create a scatterplot matrix of these
    returns, and compute the mean vector, covariance matrix, and vector
    of standard deviations of the returns.

-   Note that returns will be percentages.

    -   `prices[2:n, ]`: These are the prices from **day 2 to the last
        day** (i.e., the "current day").

    -   `prices[1:(n-1), ]`: These are the prices from **day 1 to the
        second-last day** (i.e., the "previous day").

    -   `prices[2:n, ] / prices[1:(n-1), ]`: This divides each day’s
        price by the previous day’s price to get the relative change.

        -   For example: if GM's price was \$100 yesterday and \$105
            today, this becomes 105/100 = 1.05

    -   ...`- 1`: This gives you the daily return as a decimal.

        -   Continuing our example: 1.05 - 1 = 0.05 → 5% return.

    <!-- -->

    -   `100 * (...)`: Converts the return into a **percentage**. 0.05
        becomes 5%

```{r}
dat = read.csv("Stock_Bond.csv", header = TRUE)

prices = cbind(dat$GM_AC, dat$F_AC, dat$CAT_AC, dat$UTX_AC,
               dat$MRK_AC, dat$IBM_AC)

n = dim(prices)[1]
returns = 100 * (prices[2:n, ] / prices[1:(n-1), ] - 1)

pairs(returns)

mean_vect = colMeans(returns)
cov_mat = cov(returns)
sd_vect = sqrt(diag(cov_mat))
```

## 1.1.Efficient Frontier and Optimal Portfolios under Constraints

Write an R program to find the **efficient frontier, the tangency
portfolio, and the minimum variance portfolio**, and plot on
“reward-risk space” the location of each of the six stocks, the
efficient frontier, the tangency portfolio, and the line of efficient
portfolios.

-   Use the constraints that $−0.1 ≤ w_j ≤ 0.5$ for each stock.

    -   The first constraint limits short sales but does not rule them
        out completely. The second constraint prohibits more than 50% of
        the investment in any single stock.

-   Assume that the annual risk-free rate is 3% and convert this to a
    daily rate by dividing by 365, since interest is earned on trading
    as well as non-trading days.

    -   `rf_annual = 3`: Annual risk-free rate (3%); Use 3 as Returns
        here has unit of percentages.

        Recall `returns = 100 * (prices[2:n, ] / prices[1:(n-1), ] - 1)`

The last few lines of code produce a legend. You were not expected to
include a legend on your own plot, but you can use this example in the
future when you are asked to provide a legend.

```{r}
## 0.INPUT
A = -0.1  # lower bound for weights
B = 0.5   # upper bound for weights
rf_annual = 3  # Annual risk-free rate (3%)
rf_used = rf_annual / 365  # Convert annual risk-free rate to daily rate

# Set - range of target portfolio returns
target_return_range = c(0.05, 0.08)  # lower and upper bounds for target returns
num_target_returns = 300  # Number of target returns to generate
```

```{r}
## 1.Initialization:
rf_daily = rf_used
M = length(mean_vect) # number of assets in the portfolio 
library(quadprog)

## 1.1.Constraint Matrix Setup:
Amat = cbind(rep(1, M), mean_vect, diag(1, nrow = M), -diag(1, nrow = M))

## 1.2.Portfolio return target range
muP = seq(target_return_range[1], target_return_range[2], length = num_target_returns)  # Create the range of target portfolio returns
sdP = muP  # Initially mirrors muP (will store actual standard deviations later)
weights = matrix(0, nrow = num_target_returns, ncol = M)  # Matrix to store the portfolio weights for each target return

## 2.Solve the optimization problem for each portfolio return target
for (i in 1:length(muP)) {
  bvec = c(1, muP[i], rep(A, M), rep(-B, M))  # Updated bounds for weights
  result = solve.QP(Dmat = cov_mat, dvec = rep(0, M), Amat = Amat,
                    bvec = bvec, meq = 2)
  sdP[i] = sqrt(2 * result$value)
  weights[i, ] = result$solution
}

## 3.Calculate Sharpe ratios and identify the tangency portfolio
sharpe = (muP - rf_daily) / sdP
ind = (sharpe == max(sharpe))  # locates the tangency portfolio
weights_portfolio = weights[ind, ]  # weights of the tangency portfolio

print(paste("Max Sharpe ratios:",  max(sharpe)))
print(paste("weights of tangency portfolio:", weights_portfolio))
```

-   `muP = seq(0.05, 0.08, length = 300)`: This creates a sequence of
    300 portfolio return target values ranging from 5% to 8%. These are
    the expected returns for the portfolios you will be optimizing.

```{r}

## Plot the efficient frontier
plot(sdP, muP, type = "l", xlim = c(0, 2.5), ylim = c(0, 0.1))

# Plot the risk-free asset point (rf_daily)
points(0, rf_daily, cex = 3, col = "blue", pch = "*")

## Plot the capital market line (CML)
lines(c(0, sdP[ind]), c(rf_daily, muP[ind]), col = "red", lwd = 3)
points(sdP[ind], muP[ind], col = "blue", cex = 3, pch = "*")

## Identify the minimum variance portfolio and plot it
ind2 = (sdP == min(sdP))
points(sdP[ind2], muP[ind2], col = "green", cex = 3, pch = "*")

## Plot the efficient set of portfolios above the min variance point
ind3 = (muP > muP[ind2])
lines(sdP[ind3], muP[ind3], type = "l", xlim = c(0, 0.25), ylim = c(0, 0.3), col = "cyan", lwd = 3)

## Annotate stock labels
text(sd_vect[1], mean_vect[1], "GM")
text(sd_vect[2], mean_vect[2], "F")
text(sd_vect[3], mean_vect[3], "UTX")
text(sd_vect[4], mean_vect[4], "CAT")
text(sd_vect[5], mean_vect[5], "MRK")
text(sd_vect[6], mean_vect[6], "IBM")


## Add a legend
legend("topleft", c("efficient frontier", "efficient portfolios", "tangency portfolio", "min var portfolio"),
       lty = c(1, 1, NA, NA), lwd = c(3, 3, 1, 1), pch = c("", "", "*", "*"),
       col = c("cyan", "red", "blue", "green"), pt.cex = c(1, 1, 3, 3))

```

## 1.2.Capital Allocation for Target Return (Using the Constrained Tangency Portfolio)

If an investor wants an efficient portfolio with an expected daily
return of 0.07%, how should the investor allocate his or her capital to
the six stocks and to the risk-free asset?

-   Assume that the investor wishes to use the tangency portfolio
    computed with the constraints $−0.1 ≤ w_j ≤ 0.5$, not the
    unconstrained tangency portfolio.

-   Let $\omega$ be the weight for the risk-free asset, $\mu_f$ be the
    risk-free rate, and $\mu_T$ be the expected return of the tangency
    portfolio. Then $\omega$ solves
    $0.07 = \omega\mu_f + (1 - \omega)\mu_T$.

The following continuation of the R program computes the weights for the
six stocks and the risk-free asset. The last line checks that the
weights sum to 1.

```{r}

expected_daily_return = 0.07
daily_rf_rate = 3/265

options(digits=3) # Set significant digits to display in R output to 3.
omega = (expected_daily_return- muP[ind]) / (daily_rf_rate- muP[ind])
adj_weights_port = (1-omega)*weights[ind]

print(paste("weight of the rf asset (w):", omega))
print(paste("weight assigned to tangency portfolio (1-w):", 1-omega))
print("adjusted weights for stocks in portfolio:")
print((1-omega)*weights[ind])


```

We see that the weight for the risk-free asset is 0.054 and the weights
for the six stocks are -0.08622, -0.00275, 0.31707, 0.36283, 0.30209,
0.05255. The first two stocks are sold short.

```{r}
# Check sum should be 1
omega + sum((1-omega)*weights[ind])
```

## 1.3.Black Monday in Dataset

Does this data set include Black Monday?

Yes, Black Monday was October 19, 1987 and data go from January 2, 1987
to Sept 1, 2006. Black Monday is the 202th day in the original data set
or the 201st day of returns. If you look in the spread sheet you will
see huge price declines that day. The returns that day were:

```{r}
returns[201,]
```

# 2.Portfolio Optimization and Efficient Frontier Analysis

Download monthly adjusted closing prices from Jan 2010 to Dec 2020, for
the following stocks: (Use
`tseries::get.hist.quote( ..., compression=’m’)` for monthly data.)

-   ENB.TO (Enbridge Inc.)
-   CP.TO (Canadian Pacific Railway Ltd.)
-   RCI-A.TO (Rogers Communications Inc.)
-   TD.TO (The Toronto-Dominion Bank)
-   L.TO (Loblaw Companies Ltd.)

Calculate their **net returns** and assume a **monthly interest rate**
of 0.20% throughout.

```{r}
library(tseries)
library(zoo)

tickers = c("ENB.TO","CP.TO","RCI-A.TO","TD.TO","L.TO")

N = length(tickers)
P=vector("list", N) # list for holding prices

for (i in 1:N) {
  cat("Downloading ", i, " out of ", N , "\n")
  P[[i]] = get.hist.quote(instrument = tickers[i], 
    start = as.Date("2010-01-01"), 
    end=as.Date("2020-12-01"), 
    compression="m", quote = "AdjClose", 
    retclass = "zoo", quiet = T)
}

# net returns
R = sapply(P, FUN=function(x){ as.numeric(diff(x) / lag(x, -1)) } )
colnames(R) = tickers # assign names
```

## 2.1.Two-Asset Portfolio Risk-Return Analysis Under Varying Correlation

Use only the first two stocks (ENB.TO and CP.TO). Calculate the sample
means and (individual) sample variances of their returns.

-   Consider the following hypothetical values for their correlation:
    $\rho = −1, −0.5, 0, +0.5, +1$.

    -   For each value of $\rho$, calculate their corresponding 2D
        variance-covariance matrix and plot the risk-return profiles of
        portfolios combining the two assets with weights
        $[w, (1 − w)], \;\; \forall w \in [−2, 3]$.

    -   Plot all profile curves on the same ($\mu_p,\sigma_p$)-space,
        using a different color for each value of $\rho$.

```{r}
## 1.Subset to First Two Assets
R2=R[,1:2]

## 2.Compute Mean and Standard Deviation of Each Stock’s Returns
MU = colMeans(R2)
SD = sqrt(diag(var(R2)))

### 2.1. Cov & Correlation
print(var(R2)); print(cor(R2));
print(paste("Real Correlation between Two Assets:",cor(R2)[1,2]))
```

```{r}
## 3.Scatter plot the Two Assets on Risk-Return Space
par(mfrow=c(1,1), mar=c(5, 4, 4, 2) + 0.1)
plot(SD,MU,pch=16,cex=2, col=2, xlim=c(0,.15), ylim=c(0.005,.025))
abline(v=0, lty=2); abline(h=0, lty=2) 

## 4.Define Portfolio Weights
w = seq(-2,+3,.01); W = cbind(w, 1-w)

## 5.Compute Portfolio Mean Returns
MU.p = W %*% MU

## 6. Loop Over Different Hypothetical Correlation Values
true_cor = round(cor(R2)[1,2],3)
rho=c(-1, -.5, 0, true_cor, .5, 1)
for(i in 1:5){
  COR=matrix(c(1,rho[i],rho[i],1),2,2)
  COV= COR * (SD%*%t(SD))
  SD.p = sqrt(rowSums((W %*% COV)*W)) # portfolio st.dev.
  lines(SD.p, MU.p, type='l', lwd=2, col=i); 
}

## 7.Add Points and Labels and Lengend
points(SD,MU,pch=16,cex=2, col=2); text(SD, MU, pch=16, colnames(R2), pos=c(1,3))
legend('topright', lwd=rep(2,5),col=1:5, 
       c("rho=-1","rho=-.5","rho=0",paste("True_rho=",true_cor), "rho=+.5","rho=+1")  )
```

Note that when the assets are *perfectly* correlated (either positively
or negatively), the minimum variance portfolio has variance zero.

## 2.2.Multi-Asset Efficient Frontier and Capital Market Line (Unconstrained)

Consider all 5 stocks together now, and use the sample mean and sample
variance-covariance matrix of their returns.

-   Plot the **efficient frontier** and the **capital market line** on
    the same ($\mu_p,\sigma_p$)-space and report the **tangency
    portfolio weights.** Plots the efficient frontier: the set of
    minimum-risk portfolios for different target returns.

```{r}
library(quadprog)

## 1.Setup
COV=cov(R)
MU=colMeans(R)
SD=sqrt(diag(COV)) # Standard deviation (risk) of each asset
N=dim(R)[2] # Number of assets 

## 1.1.Setup for Efficient Frontier
Amat = cbind(rep(1,N),MU)
mu.p = seq( -.005, .05,length=100) # Target returns to build efficient frontier
sd.p = mu.p # Placeholder for standard deviations 
mu.f = .002 # monthly risk-free interest rate

```

-   `Amat` holds the constraint matrix:

    -   First column (`rep(1,N)`): sum of weights = 1

    -   Second column (`MU`): expected return equals the target return

-   For each target return `mu.p[i]`, solve the following **optimization
    problem**:

$$ \min_{w} \frac{1}{2} w^T \Sigma w \space \space \space\space subject \space to : \space \sum w_i = 1, \quad \mu^T w = \mu_p $$

-   `W.tang=COV.i%%(MU-mu.f) / sum( COV.i%%(MU-mu.f) )`

    -   This computes the **tangency portfolio** (risky portfolio that
        maximizes the Sharpe ratio):

    $$ w^* = \frac{\sum^{-1} (\mu - r_f)}{1^T \sum^{-1} (\mu - r_f)} $$

```{r}
## 2.Plot Assets in Risk-Return Space
plot(SD, MU, pch=16, cex=1.2, col= 2, xlim=c(0,.1), ylim=c(0,.025))
abline(v=0, lty=2); abline(h=0, lty=2) 
text(SD, MU, tickers, cex=1, pos=4) # Labels points with its ticker.

## 3.Compute Efficient Frontier via Quadratic Programming
for (i in 1:length(mu.p)) 
{
  bvec=c(1,mu.p[i]) 
  out=solve.QP(Dmat=2*COV,dvec=rep(0,N),Amat=Amat,bvec=bvec,meq=2)
  sd.p[i] = sqrt(out$value) # Minimum portfolio variance(objective function value)
}

lines(sd.p,mu.p,type="l", lwd=2, col=2) # plot least variance portfolios

## 4.Compute Tangency Portfolio (Maximum Sharpe Ratio)
COV.i=solve(COV) #inverse of the matrix COV
W.tang=COV.i%*%(MU-mu.f) / sum( COV.i%*%(MU-mu.f) ) 
mu.tang=sum(W.tang*MU); sd.tang=sqrt(sum( (COV %*% W.tang) * W.tang ) )

## 5.Plot the Capital Market Line (CML)
points( sd.tang, mu.tang, pch=15, cex=1.3, col=2)  
sharpe=(mu.tang-mu.f)/sd.tang
abline(mu.f,sharpe,lwd=2,lty=2,col=2)
```

Aside: if want to calculate
$$ w^{**} = \frac{\sum^{-1} 1}{1^T \sum^{-1} 1} $$

```{r}
Sigma = var(R) # or cov(R)
Sigma_inv = solve(Sigma)
w_mv_starstar = rowSums(Sigma_inv)/sum(Sigma_inv)
w_mv_starstar
```

-   **CML**: Capital Market Line that starts at the risk-free rate and
    goes through the tangency portfolio, with slope equal to the Sharpe
    ratio: $$\text{CML: } \mu = r_f + \text{Sharpe} \cdot \sigma$$

```{r}
print(paste("The Sharpe ratio is:",sharpe))
```

The tangency portfolio weights are:

```{r}
round( t(W.tang), 4 )
```

## 2.3.EF,CML with Constraints

Repeat the previous part b. (i.e. plot the **efficient frontier** and
**capital market line,** and report the **tangency portfolio weights**)
with the restriction.

-   all weights are within the bounds:
    $0 \le w_i \le 0.5,\;\; \forall i = 1, \ldots, 5$.

```{r}
## 1.INPUT - Set Up
lwb = 0 # lwb <= weights
upb = 0.5 # weights<= upb
mu.pot = seq( min(MU), max(MU),length=300) # potential mean returns


## 2.Plot Individual Assets
plot(SD, MU, pch=16, cex=1.2, col= 2,  xlim=c(0,.1), ylim=c(.0,.025))
abline(v=0, lty=2); abline(h=0, lty=2);text(SD, MU, tickers, cex=1, pos=4)

## 3.Initializations
Amat = cbind(rep(1,N),MU,diag(1,nrow=N),-diag(1,nrow=N))
mu.p = NULL # initialize portofio standard error 
sd.p = NULL # initialize portofio standard error 
W.p = NULL # initialize portofio weights 

## 4.Loop to Solve QP for Each Target Return
for (i in 1:length(mu.pot)) 
{bvec=c(1,mu.pot[i],rep(lwb,N),rep(-upb,N)) # Updated bounds for weights
  #check whether potential mean return can be achieved with given constraints
  out=tryCatch( solve.QP(Dmat=2*COV,dvec=rep(0,N),Amat=Amat,bvec=bvec,meq=2), error=function(e) NULL) 
  #if mean return is achievable, save its st.dev. & portfolio weights
  if(!is.null(out)){ #  if
    mu.p=c(mu.p, mu.pot[i]); sd.p=c(sd.p, sqrt(out$value))
    W.p=rbind(W.p, out$solution)}}

## 5.Plot the Constrained Efficient Frontier
lines(sd.p,mu.p,type="l", lwd=2, col=2) # plot least variance portfolios
colnames(W.p)=tickers

## 6.Compute Tangency Portfolio (Max Sharpe Ratio) Under Constraints
sharpe=( mu.p-mu.f)/sd.p; ind.tang=which.max(sharpe)
W.tang=W.p[ind.tang,] # weights of tangency portfolio
sd.tang = sd.p[ind.tang] # st.dev of tangency portfolio
mu.tang = mu.p[ind.tang] # expected return of tangency portfolio

## 7.Plot Tangency Portfolio and CML
points( sd.tang, mu.tang, pch=15, cex=1.3, col=2)  
abline( c(mu.f, sharpe[ind.tang]),lwd=2, col=2, lty=2)
```

-   The set of feasible portfolios will be a subset of that of the
    unconstrained problem. Note that the *constrained* efficient
    frontier is not a parabola any more.

-   But the constrained tangency portfolio is the same as the
    unconstrained one.

    -   Since the unconstrained tangency portfolio weights were all
        within [0,0.5], and have (approximately) the same Sharpe ratio
        `r max(sharpe)`.
    -   Note that the constrained tangency portfolio weights are
        (approximately) the same as the unconstrained ones:

```{r}
max(sharpe)
```

```{r}
round( W.tang, 4 )
```

## 2.4.EF,CML with Constraints ( Not Same Weights)

what if we changed the constraint, to -0.111\<=w1 \<= 0.5101,
-0.202\<=w2 \<= 0.6201, -1=\<w3\<=1, -0.4\<=w4 \<= 0.8004, -1\<=w5\<=1

```{r}
## 1.INPUT - Set individual bounds for each asset
lwb = c(-0.111, -0.202, -1, -0.4, -1)   # lower bounds for w1 to w5
upb = c(0.5101, 0.6201,  1, 0.8004, 1)  # upper bounds for w1 to w5
mu.pot = seq( min(MU), max(MU),length=300) # potential mean returns


## 2.Plot Individual Assets
plot(SD, MU, pch=16, cex=1.2, col= 2,  xlim=c(0,.1), ylim=c(.0,.025))
abline(v=0, lty=2); abline(h=0, lty=2);text(SD, MU, tickers, cex=1, pos=4)

## 3.Initializations
Amat = cbind(rep(1,N),MU,diag(1,nrow=N),-diag(1,nrow=N)) # here 1 for sum(w)=1, MU for expected return, others for bounds
mu.p = NULL # initialize portofio standard error 
sd.p = NULL # initialize portofio standard error 
W.p = NULL # initialize portofio weights 

## 4.Loop to Solve QP for Each Target Return
for (i in 1:length(mu.pot)) {
  bvec = c(1, mu.pot[i], lwb, -upb)  # include your custom bounds here
  out = tryCatch(solve.QP(Dmat=2*COV, dvec=rep(0,N), Amat=Amat, bvec=bvec, meq=2),
    error=function(e) NULL)
  if (!is.null(out)) {
    mu.p = c(mu.p, mu.pot[i]); sd.p = c(sd.p, sqrt(out$value))
    W.p = rbind(W.p, out$solution)}}

## 5.Plot the Constrained Efficient Frontier
lines(sd.p,mu.p,type="l", lwd=2, col=2) # plot least variance portfolios
colnames(W.p)=tickers

## 6.Compute Tangency Portfolio (Max Sharpe Ratio) Under Constraints
sharpe=( mu.p-mu.f)/sd.p; ind.tang=which.max(sharpe)
W.tang=W.p[ind.tang,] # weights of tangency portfolio
sd.tang = sd.p[ind.tang] # st.dev of tangency portfolio
mu.tang = mu.p[ind.tang] # expected return of tangency portfolio

## 7.Plot Tangency Portfolio and CML
points( sd.tang, mu.tang, pch=15, cex=1.3, col=2)  
abline( c(mu.f, sharpe[ind.tang]),lwd=2, col=2, lty=2)
```

```{r}
print(paste("The Max Sharpe ratio is:",max(sharpe)))
print("constrained tangency portfolio weights:")
print(round( W.tang, 4 ))
```
