---
title: "Factor Models"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Regression - Mutual Funds Sector Attribution and Performance Analysis

For this problem you will use regression to identify the composition of
various mutual funds.

## 1.1.Data and Return for Mutual Funds

Download the adjusted daily closing prices from Jan 1 2020 to Dec 31
2022 for the 5 mutual funds below (use `tseries::get.hist.quote()` for
each ticker):

-   FCNTX: Fidelity Contrafund

-   PIODX: Pioneer A

-   AIVSX: American Funds Invmt Co of Amer A

-   PRBLX: Parnassus Core Equity Investor

-   VFIAX: Vanguard 500 Index Admiral

Note that each of these funds has at least 90% of their weight in the US
stocks market. You can actually check the composition of the investment
over different stock sectors from Yahoo Finance, under the fund’s
holdings tab; e.g. for FCNTX at
<https://finance.yahoo.com/quote/FCNTX/holdings>.

```{r,warning=FALSE}
library(zoo)
library(tseries)

MF.names=c('FCNTX','PIODX','AIVSX','PRBLX','VFIAX')
N.MF=length(MF.names)
S=list()
for(i in 1:N.MF){
  S[[i]]=get.hist.quote(MF.names[i], start='2022-01-01', end='2022-12-31', quote='AdjClose', quiet = TRUE)
}
R=lapply(S, FUN = function(x){ diff(x) /lag(x,-1) }) # MF net returns
logReturn = lapply(S, FUN = function(x){ diff(log(x)) })
RY=matrix(unlist(R),ncol=N.MF) # bind returns in a matrix
colnames(RY)=MF.names
```

-   **`lapply(...)`**: Applies the function to each time series of
    prices in list **`S`**.

-   **`diff(x)`**: Calculates the difference between consecutive days
    (i.e., $P_t$ - $P_{t-1}$).

-   **`lag(x, -1)`**: Shifts the prices forward by one time unit, i.e.
    $P_{t-1}$

-   MF net returns: $Return_t = \frac{P_t - P_{t-1}}{P_{t-1}}$

-   `diff(log(x))`: Computes the difference of log prices, which gives
    you:

    $LogReturn_{t} = r_t = log⁡(P_{t})−log⁡(P_{t−1})=\log\left(\frac{P_t}{P_{t-1}}\right) = log(1+R_{t})$

## 1.2.Regression - Sector Attribution on Sector ETFs

Assume you do not have any information about the investment strategy of
the funds. Download the daily prices and calculate returns of the
following EFTs, which track different sectors of the economy:

-   XLB: Basic Materials \| XLY: Consumer Cyclical \| XLF: Financial
    Services \| VNQ: Real Estate \|

    XLP: Consumer Defensive \| XLV: Healthcare \| XLU: Utilities \| XTL:
    Communication Services \|

    XLE: Energy \| XLI: Industrials \| XLK: Technology

Regress each of the mutual fund returns on the above ETF returns and
create barplots of the estimated beta coefficients. Do these accurately
reflect the allocation over the different sectors (as described in Yahoo
Finance)?

```{r}
ETF.names=c('XLB','XLY','XLF','VNQ','XLP','XLV','XLU','XTL','XLE','XLI','XLK')
sectors = c("Basic Materials","Consumer Cyclical","Financial Services",
            "Real Estate","Consumer Defensive", "Healthcare", "Utilities", 
            "Communication Services", "Energy", "Industrials", "Technology")
N.ETF=length(ETF.names)
S=list()
for(i in 1:N.ETF){
  S[[i]]=get.hist.quote(ETF.names[i], start='2022-01-01',                        end='2022-12-31', quote='AdjClose', quiet = TRUE)
}
R=lapply(S, FUN = function(x){ diff(x) / lag(x,-1) }) # calculate ETF returns
logReturn = lapply(S, FUN = function(x){ diff(log(x)) })
RX=matrix(unlist(R),ncol=N.ETF) # bind returns in a matrix
colnames(RX)=ETF.names

out=list(); 
for(i in 1:N.MF){
  out[[i]]=lm(RY[,i] ~ RX)
  weights = out[[i]]$coef[-1] 
  barplot( rev(weights), names.arg = rev(sectors), main = MF.names[i],
          horiz = TRUE, las = 2, cex.names = .8)
}
```

The barplots of the regression coefficients (betas) **roughly** follow
the sector weightings for each fund. Nevertheless, they are not always
close in actual value (e.g. in some cases the betas are negative, even
though weightings are positive). The differences can be due to the fact
that we use ETFs as *proxies* for a sector, but the actual holding of
the fund within the sector might be different. Moreover, there will be
estimation error in our regression model, which is only based on the
last year's returns.

Nevertheless, it is quite impressing that we can (approximately)
identify the strategy of a fund, without knowing anything beyond its
past returns. This approach works because of the linear formula for net
portfolio returns: $$ R_p = w_1 R_1 + \cdots + w_N R_N$$ Regressing
portfolio returns on other assets, we can estimate the weights, assuming
the portfolio composition is constant.

## 1.3.Jensen's Alpha and Fund Performance

Compare the performance of the mutual funds to that of a portfolio of
ETFs by reporting the value of **Jensen’s alpha** (based on the
regressions from the previous part) and its corresponding **p-value**.

```{r}
alpha=p.val=rep(0,N.MF)
for(i in 1:N.MF){
  alpha[i]=out[[i]]$coefficients[1]*250 # annualized Jensen alpha
  p.val[i]=summary(out[[i]])$coefficients[1,4]
}
cbind(alpha, p.val)
```

All the funds' alphas are negative, although their p-values are not very
small. A likely cause for this is that funds charge a *fee* which
consistently eats up some of the returns of their constituent assets.
ETFs have typically lower fees than mutual funds, but our regression
does not account for **transaction costs** (it is more costly to buy
multiple assets than a single one), so the comparison is more nuanced.

Note that you can find Jensen alphas and other performance measures
(e.g., Sharpe & Treynor Ratios) for assets in Yahoo! Finance under the
[risk tab](https://finance.yahoo.com/quote/FCNTX/risk?p=FCNTX). These
metrics are based on the CAPM/Market factor model, by regressing the
asset's returns on a proxy for the market return (e.g., S&P500).

# 2.Factor Analysis and Simulation

Use the following R code to download daily prices of 10 ETFs, from Jan
1, 2018 to Dec 31, 2019.

```{r}
library(zoo)
library(tseries)
tickers = c("DVEM", "EXT", "HYEM", "LTPZ", "SCHP", 
            "EDV",  "SPMB", "TLT", "GOVT")
S=list()
for(i in 1:length(tickers)){
  S[[i]] = get.hist.quote(tickers[i], start='2018-01-01', end='2019-12-31', quote='AdjClose', drop = TRUE)  
}

```

## 2.1.Return, LogReturn and plot

Calculate the log-returns of the ETFs, and plot the price and return
series for the first ETF (DVEM)

```{r}
r = lapply(S, FUN=function(x){diff(log(x))}) 
R = lapply(r, FUN=function(x){exp(x)-1}) 

plot(S[[1]])
plot(R[[1]])

```

## 2.2.Fact Analysis

Use `factanal()` to fit a 2-factor model to the correlation matrix of
the returns. Report the factor loadings and idiosyncratic variances of
you model.

```{r}

Rmat = simplify2array(R) # Simplify list of matrices into a single array
fmod = factanal( Rmat, factors = 2,  lower = 0.005)

( b = fmod$loadings )
( v = fmod$uniquenesses )
```

Or can do :

```{r}

RX=matrix(unlist(R),ncol=length(tickers))
fact = factanal(RX, factors = 2, rotation = "none")
#print(fact)
```

## 2.3.Multivariate normal simulation using Factor-model correlations

Simulate 250 daily log-returns using a **multivariate Normal
distribution** with parameters given by the sample means and variances
of the ETFs, and **correlation matrix given by the previous factor
model**. Calculate and plot the cumulative net-returns of an equally
weighted portfolio over the 10 ETFs.

```{r}
MU = sapply(R, mean)
SD = sapply(R, sd)
VC = ( b%*%t(b) + diag(v) ) * (SD %*% t(SD)) 

library(mvtnorm)
Rsim = mvtnorm::rmvnorm(250, MU, VC)

R_eqwt = rowMeans( exp( apply(Rsim,2,cumsum) ) - 1 )
plot(R_eqwt, type = "l");
```

# 3.Factor Models of Asset Returns

In this section, we will start with the **one-factor CAPM model** of
Chap. 17 and then extend this model to the **three-factor Fama–French
model**.

-   We will use the dataset Stock_Bond_2004_to_2005.csv on the book’s
    website, which contains stock prices and other financial time series
    for the years 2004 and 2005. Data on the Fama–French factors are
    available at Prof. Kenneth French’s website, where `RF` is the
    **risk-free rate** and `Mkt.RF`, `SMB`,and `HML` are the
    **Fama–French factors.**

Go to Prof. French’s website and get the daily values of `RF`, `Mkt.RF`,
`SMB`, and `HML` for the years 2004–2005. It is assumed here that you’ve
put the data in a text file FamaFrenchDaily.txt. Returns on this website
are expressed as **percentages**.

## Model 1: One-Factor CAPM

Now fit the CAPM to the four stocks using the `lm` command. This code
fits a linear regression model separately to the four responses. In each
case, the independent variable is `Mkt.RF`.

We apply a linear regression model where excess returns of four selected
stocks (GM, Ford, UTX, Merck) are regressed on the market excess return
(`Mkt.RF`).

```{r,warning=FALSE}
# Uses daily data 2004-2005
stocks = read.csv("Stock_FX_Bond_2004_to_2005.csv",header=T)
attach(stocks)

FF_data = read.table("FamaFrenchDaily.txt", header = TRUE)
FF_data = FF_data[-1, ] # delete first row since stocks_diff

stocks_subset = as.data.frame(cbind(GM_AC, F_AC, UTX_AC, MRK_AC))
# Calculating Log Returns and Differencing
stocks_diff = as.data.frame(100 * apply(log(stocks_subset),2, diff)- FF_data$RF)
names(stocks_diff) = c("GM", "Ford", "UTX", "Merck")


# lost a row due to differencing
fit1 = lm(as.matrix(stocks_diff) ~ FF_data$Mkt.RF)
summary(fit1)
```

### (a).Testing Intercepts

The CAPM predicts that all four intercepts will be zero. For each stock,
using α =0.025, can you accept the null hypothesis that its intercept is
zero? Why or why not? Include the p-values with your work.

-   The intercepts p-value is below 0.025 for General motors (GM) and
    Ford, but not for United Technologies Incorporatted (UTX) and Merck.
    Of course, a p-value only shows statistical significance, not the
    size of an effect.

-   However, estimated intercepts for GM and Ford are -0.23 and -0.18
    and these are reasonably large in magnitude. Since they are
    negative, this suggests than these two stocks were overpriced.

### (b).Residual Correlations

The CAPM also predicts that the four sets of residuals will be
uncorrelated. What is the correlation matrix of the residuals? Give a
95% confidence interval for each of the six correlations. Can you accept
the hypothesis that all six correlations are zero?

-   The correlation matrix is below. All correlations are reasonably
    close to 0 (less than 0.1 in magnitude) except the correlation
    between GM and Ford residuals.

-   That correlation is 0.52 and has a very small p-value. The
    correlation between GM and Merck residuals is -0.0878 and is
    statistically significant at 0.05 but might be too small to be of
    practical significance.

```{r}
cor(residuals(fit1)) 
res = residuals(fit1)
cor.test(res[,"GM"],res[,"Ford"])
cor.test(res[,"GM"],res[,"UTX"])
cor.test(res[,"GM"],res[,"Merck"])
cor.test(res[,"Ford"],res[,"UTX"])
cor.test(res[,"Ford"],res[,"Merck"])
cor.test(res[,"UTX"],res[,"Merck"])

```

### (c).Covariance Matrix CAPM Estimation

Regardless of your answer to Problem 6, assume for now that the
residuals are **uncorrelated**. Then use the CAPM to estimate the
**covariance matrix** of the excess returns on the four stocks. Compare
this estimate with the **sample covariance matrix** of the excess
returns. Do you see any large discrepancies between the two estimates of
the covariance matrix?

-   We see below that the estimated covariance matrix using the CAPM is
    similar to the sample covariance matrix, with the exception of the
    covariance between GM and Ford. Since these two stocks have a high
    residual correlation and the CAPM assumes that the residual
    correlation is 0, it is not surprising that the CAPM estimated
    covariance matrix severely underestimates the correlation between GM
    and Ford.

```{r}
attach(FF_data)
sigF = var(Mkt.RF)
bbeta = as.matrix(fit1$coef)
bbeta = bbeta[-1,] # delete intercepts so bbeta has the four slopes
n=dim(stocks_diff)[1]
sigeps = as.matrix((var(as.matrix(res))))

sigeps_ind = diag(as.matrix(sigeps))
sigeps_ind = diag(sigeps_ind,nrow=4)
cov_equities = sigF* bbeta %*% t(bbeta) + sigeps_ind
cov_equities
cov(stocks_diff)

```

If use the true residual correlation.

```{r}
bbeta_mat <- matrix(bbeta, nrow = 1)
cov_equities = t(bbeta_mat) %*% sigF %*% (bbeta_mat) + sigeps_ind
cov_equities
cov(stocks_diff)

```

------------------------------------------------------------------------

## Model 2: Fama–French Three-Factor Model

Next, you will fit the Fama–French three-factor model. Run the following
R code, which is much like the previous code except that the regression
model has two additional predictor variables, SMB and HML.

```{r}
fit2 = lm(as.matrix(stocks_diff) ~ FF_data$Mkt.RF +
FF_data$SMB + FF_data$HML)
summary(fit2)

```

### (a).Testing Additional Factor Relevance

The CAPM predicts that for each stock, the slope (beta) for SMB and HML
will be zero. Explain why the CAPM makes this prediction. Do you accept
this null hypothesis? Why or why not?

(NO SOLUTION)

### (b).Residual Correlations

If the Fama–French model explains all covariances between the returns,
then the correlation matrix of the residuals should be diagonal. What is
the estimated correlations matrix? Would you accept the hypothesis that
the correlations are all zero?

(NO SOLUTION)

```{r}
cor(residuals(fit2)) 
res = residuals(fit2)
cor.test(res[,"GM"],res[,"Ford"])
cor.test(res[,"GM"],res[,"UTX"])
# cor.test(res[,"GM"],res[,"Merck"])
# cor.test(res[,"Ford"],res[,"UTX"])
# cor.test(res[,"Ford"],res[,"Merck"])
# cor.test(res[,"UTX"],res[,"Merck"])
```

### (c).Model Comparison CAPM vs Fama-French

Which model, CAPM or Fama–French, has the smaller value (better) of AIC?
Which has the smaller value of BIC? What do you conclude from this?

(NO SOLUTION)

The `AIC()` and `BIC()` functions need models with a **single response**
(like just `y ~ x`). For multivariate models, `logLik()` (which AIC/BIC
rely on) isn't defined by default.

```{r}
fit_AICBIC=fit1
# Residual Sum of Squares
rss <- sum(residuals(fit_AICBIC)^2)

# Number of observations and parameters
n=dim(stocks_diff)[1]
k <- length(coef(fit_AICBIC))  # or use length(fit$coefficients) * number of responses

# AIC and BIC (based on Gaussian likelihood approximation)
(aic <- n * log(rss / n) + 2 * k)
(bic <- n * log(rss / n) + log(n) * k)

```

### (d).Covariance Matrix

#### Factor Covariance Matrix

What is the covariance matrix of the three Fama–French factors?

```{r}
#attach(FF_data)
SIGMA_F = cov(FF_data[,c(2,3,4)])
SIGMA_F
```

#### Covariance Matrix of Return - by Fama–French Model

```{r}
fit2 = lm(as.matrix(stocks_diff) ~ FF_data$Mkt.RF+FF_data$SMB + FF_data$HML)
# summary(fit2)

```

```{r}
res = residuals(fit2)

FF3=FF_data[,c(2,3,4)]
SIGMA_F = cov(FF3)
bbeta = as.matrix(fit2$coef)
bbeta = bbeta[-1,] # delete intercepts so bbeta has the four slopes
n=dim(stocks_diff)[1]
sigeps = as.matrix((var(as.matrix(res))))
sigeps_ind = diag(as.matrix(sigeps))
sigeps_ind = diag(sigeps_ind,nrow=4)
cov_equities_indeps = t(bbeta) %*% SIGMA_F %*% (bbeta) + sigeps_ind

cov_equities = t(bbeta) %*% SIGMA_F %*%(bbeta) + sigeps
cov_equities_indeps;cov_equities;cov(stocks_diff)
```

### (e).Predicting Variance and Covariance for New Stocks

In this problem, Stocks 1 and 2 are two stocks, not necessarily in the
Stock_FX_Bond_2004_to_2005.csv data set. Suppose that Stock 1 has betas
of 0.5,0.4,and −0.1 with respect to the three factors in the Fama–French
model and a residual variance of 23.0. Suppose also that Stock 2 has
betas of 0.6, 0.15, and 0.7 with respect to the three factors and a
residual variance of 37.0. Regardless of your answer to Problem 9, when
doing this problem, assume that the three factors do account for all
covariances.

(a) Use the Fama–French model to estimate the variance of the excess
    return on Stock 1.

    ```{r}
    # factor covariance matrix
    Sigma_F <- cov(FF_data[,3:5],FF_data[,3:5])  # columns: Mkt.RF, SMB, HML


    # Define betas and residual variance
    beta1 <- c(0.5, 0.4, -0.1)
    resid_var1 <- 23.0


    # Calculate the factor-related variance
    var_factor <- t(beta1) %*% Sigma_F %*% beta1

    # Total variance
    (var_stock1 <- var_factor + resid_var1)

    ```

(b) Use the Fama–French model to estimate the variance of the excess
    return on Stock 2. And use the Fama–French model to estimate the
    covariance between the excess returns on Stock 1 and Stock 2.

    ```{r}
    # Define betas for Stock 2 and residual variance
    beta2 <- c(0.6, 0.15, 0.7)
    resid_var2 <- 37.0

    # Calculate the variance for Stock 2
    var_factor2 <- t(beta2) %*% Sigma_F %*% beta2
    (var_stock2 <- var_factor2 + resid_var2)

    # Calculate the covariance between Stock 1 and Stock 2
    (cov12 <- t(beta1) %*% Sigma_F %*% beta2)

    ```

(NO SOLUTION)

------------------------------------------------------------------------

## Model 3: Statistical Factor Analysis

This section applies statistical factor analysis to the log returns of
10 stocks in the data set Stock_FX_Bond.csv. The data set contains
adjusted closing (AC) prices of the stocks, as well as daily volumes and
other information that we will not use here.

### (a).Exploratory Factor Modeling

The following R code will read the data, compute the log returns, and
fit a two-factor model. Note that `factanal` works with the correlation
matrix or, equivalently, with standardized variables.

```{r}
dat = read.csv("Stock_FX_Bond.csv")
stocks_ac = dat[ , c(3, 5, 7, 9, 11, 13, 15, 17)]
n = length(stocks_ac[ , 1])
stocks_returns = log(stocks_ac[-1, ] / stocks_ac[-n, ])

fact = factanal(stocks_returns, factors = 2, rotation = "none")
print(fact)
```

Loadings less than the parameter cutoff are not printed. The default
value of cutoff is 0.1, but you can change it as in
`print(fact,cutoff = 0.01)` or `print(fact, cutoff = 0)`.

### (b).Interpreting Factor Loadings

What are the factor loadings? What are the variances of the unique risks
for Ford and General Motors?

-   The factor loadings and uniqueness are in the output below. We see
    that the uniquenesses for Ford and GM are 0.423 and 0.399,
    respectively.

### (c).Number of Factors Needed

Does the likelihood ratio test suggest that two factors are enough? If
not, what is the minimum number of factors that seems sufficient?

-   The results of the likelihood ratio tests below strongly suggest
    that there are more than two factors. It seems that 4 factors are
    sufficient, but not 3.

```         
Test of the hypothesis that 2 factors are sufficient.
The chi square statistic is 564.66 on 13 degrees of freedom.
The p-value is 2.6e-112

Test of the hypothesis that 3 factors are sufficient.
The chi square statistic is 162.29 on 7 degrees of freedom.
The p-value is 1.06e-31

Test of the hypothesis that 4 factors are sufficient.
The chi square statistic is 0.3 on 2 degrees of freedom.
The p-value is 0.86 
```

### (d).Estimating Correlations from Factor Model

Regardless of your answer to Problem, use the two-factor model to
estimate the correlation of the log returns for Ford and IBM.

The output below contains the **estimated correlation matrix using the
factor model** after line 4. For comparison, the **sample correlation
matrix** is also printed.

```{r}
loadings = matrix(as.numeric(loadings(fact)), ncol = 2)
unique = as.numeric(fact$unique)
options(digits=2)
loadings %*% t(loadings) + diag(unique)
cor(stocks_returns)
```
