---
title: "Risk Management"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Risk Measures

## Data

Get last 23years of daily IBM stock prices

```{r,warning=FALSE}
library(tseries)
library(zoo)

start.date=as.Date("1990-01-01")

# Get last 23years of daily IBM stock prices
Padj=get.hist.quote(instrument = "IBM", start = start.date , quote = "AdjClose", retclass = "zoo", quiet = T)
plot(Padj)

# R=diff(Padj)/as.vector(Padj[-length(Padj)]) # net returns
R = diff(Padj)/lag(Padj,k=-1)  # net returns
r=log(1+R) # Log returns. r = log( R + 1 ) <=> R = exp( r ) - 1
plot(R)
```

## 1.1.Parametric VaR using t-distr

```{r,warning=FALSE}
library(MASS) # for fitdistr()
parm=fitdistr(R, "t") # get Student's t MLE of (mu,sigma,nu)
mu=parm$est[1]; sig=parm$est[2]; dF=parm$est[3]
alpha=.01

t.quant=qt(alpha,dF,lower.tail = TRUE)
(VaR.t= -mu -sig * t.quant)
(CVaR.t= -mu + sig/alpha * dt(t.quant,dF) * (dF+t.quant^2)/(dF-1) )

```

```{r,warning=FALSE}
library(MASS) # for fitdistr()
L = -R
parm=fitdistr(L, "t") # get Student's t MLE of (mu,sigma,nu)
mu=parm$est[1]; sig=parm$est[2]; dF=parm$est[3]
alpha=.01

t.quant=qt(alpha,dF,lower.tail = FALSE)
(VaR.t= mu+sig * t.quant)
(CVaR.t= mu + sig/alpha * dt(t.quant,dF) * (dF+t.quant^2)/(dF-1) )

```

## 1.2.Historical VaR

```{r}

alpha=.01

(VaR.hist=-quantile(R,alpha))
(CVaR.hist=-mean(R[which(R<(-VaR.hist))]))

```

```{r}


plot(R); 
abline(h=-VaR.t, col=2, lty=5, lwd=1)
abline(h=-CVaR.t, col=2, lty=3, lwd=1)

abline(h=-VaR.hist, col=3, lty=5, lwd=1)
abline(h=-CVaR.hist, col=3, lty=3, lwd=1)

plot(density(R))
abline(v=-VaR.t, col=2, lty=5, lwd=1)
abline(v=-CVaR.t, col=2, lty=3, lwd=1)

abline(v=-VaR.hist, col=3, lty=5, lwd=1)
abline(v=-CVaR.hist, col=3, lty=3, lwd=1)

```

If use Loss directly

```{r}
alpha=.01
L=-R
(VaR.hist=quantile(L,1-alpha))
(CVaR.hist=mean(L[which(L>(VaR.hist))]))
```

### 1.2.1.VaR historical

Calculate the historical Value-at-Risk (VaR) at 95% confidence level.
According to this VaR, if you invest \$1,000 in stock, the value of your
investment after 1 day would exceed which amount with 95% confidence?
(Hint: the historical VaR is minus the 5% sample quantile of the
returns; use the quantile function in R.

```{r}
CI = 0.95 ; a = 1-CI # CI=1-a
VaR = - quantile( as.numeric(R), probs = a)
# with 95% confidence, investment exceeds:
1000 * (1-VaR)
```

### 1.2.2.CVaR historical

Calculate the historical Conditional Value-at-Risk (CVaR). According to
this CVaR, if you invest \$1,000 in stock, what would be your expected
1-day loss, conditional on the 5% worst cases?

```{r}
CVaR = -mean( as.numeric(R)[ R < -VaR ] )
# expected loss in 5% worst case
1000*CVaR
```

## 1.3.Time Series - RiskMetrics Model

Simple time series model for return volatility (P29 -30 lecture W6)

```{r}
V2=var(R); lambda=.94
n=length(R); alpha=.01

for(i in 2:n){
  V2[i]=lambda*V2[i-1]+(1-lambda)*R[i-1]^2 
}
V2=zoo(V2,index(R)) # fitted RiskMetrics volatilities square
VaRa = sqrt(V2)*qnorm(1-alpha)
CVaRa = sqrt(V2)*dnorm(qnorm(alpha))/alpha

plot(R)
lines(-VaRa, col=3) # lines(sqrt(V2)*qnorm(alpha), col=2)
lines(-CVaRa, col=4) 
```

## 1.4.Time Series - ARCH/GARCH models

Garch(p,q) models.

```{r}
library(fGarch)
V2=var(R);alpha=.01

fit=garchFit( ~garch(1,1), data = R, trace = F) # fit GARCH(1,1) model with Normal errors
summary(fit)

V=zoo(fit@sigma.t,index(R)) # fitted GARCH volatilities, sigma
VaRa = V*qnorm(1-alpha)
CVaRa = V*dnorm(qnorm(alpha))/alpha
plot(R)
lines(sqrt(V2)*qnorm(alpha), col=2)
lines(-VaRa, col=3) # 
lines(-CVaRa, col=4) # 
```

# 2.Univariate t - VaR and ES

In this section we will compare VaR and ES parametric (unconditional)
estimates with those from using ARMA+GARCH (conditional) models.
Consider the daily returns for Coca-Cola stock from January 2007 to
November 2012.

Consider the daily returns for Coca-Cola stock from January 2007 to
November 2012.

```{r}
CokePepsi = read.table("CokePepsi.csv", header=T)
price = CokePepsi[,1]
returns = diff(price)/lag(price)[-1] # Net Return
ts.plot(returns)
```

Assume that the returns are iid and follow a t-distribution. Run the
following commands to get parameter estimates in R.

```{r}
S = 4000
alpha = 0.05
library(MASS)
res = fitdistr(returns,'t')
mu_return = res$estimate['m']
lambda = res$estimate['s'] # Sigma
nu = res$estimate['df']
qt(alpha, df=nu)
dt(qt(alpha, df=nu), df=nu)
```

For an investment of \$4,000, what are estimates of $VaR^t(0.05)$ and
$ES^t(0.05)$?

$VaR^t(0.05)$ is 75.31 and $ES^t(0.05)$ is 122.1. See the output below.

```{r}

mu_loss=-mu_return

VaR =S*( mu_loss + lambda*qt(1-alpha,df=nu))
options(digits=4);VaR

den = dt(qt(alpha, df=nu), df=nu)
ES = S * (mu_loss + lambda*(den/alpha)* (nu+qt(alpha, df=nu)^2 )/(nu-1))
ES

```

# 3.EVaR

Consider the example with the two risky zero-coupon bonds priced at \$95
per \$100 face value, where each has 4% default probability
independently of the other.

## 3.1.Entropic Value-at-Risk (EVaR) by optimize()

Calculate the $\alpha = 5\%$ Entropic Value-at-Risk (EVaR) for one of
these bonds. Use numeric minimization: `optimize()` in R, to find EVaR.

------------------------------------------------------------------------

The (marginal) loss distribution of each bond ($L_{1/2}$) is given by
the PMF

$$p_L(\ell) = \mathbb{P}(L = \ell) = 
\begin{cases} 
0.04, & \ell = 95 - 0 = 95 \\ 
0.96, & \ell = 95 - 100 = -5 
\end{cases}$$

with MGF

$$M_L(z) = \mathbb{E}[e^{zL}] = 0.04e^{95z} + 0.96e^{-5z}$$

The EVaR at $\alpha$ is given by

$$EVaR_\alpha = \inf_{z>0}\left\{\frac{\ln(M_L(z)/\alpha)}{z}\right\} = \inf_{z>0}\left\{\frac{\ln\left((0.04e^{95z} + 0.96e^{-5z})/0.05\right)}{z}\right\}$$

```{r}
fn = function(z){ log((0.04*exp(95*z)+0.96*exp(-5*z))/0.05)/z }
optimise(fn, c(0,1))
```

Running this minimization w.r.t. $z$ in R, we get that the minimum is
$EVaR_{0.05}(L)=92.10402$, occurring at $z=0.06690106$.

## 3.2.Subadditive Verify

Calculate the EVaR of a portfolio of two of these bonds, and show that
it is subadditive.

The loss distribution for the sum of the two bonds $(L_1 + L_2)$ is

$$p_{L_1 + L_2}(\ell) = \mathbb{P}(L_1 + L_2 = \ell) = 
\begin{cases} 
(0.04)^2 = 0.0016, & \ell = 95 + 95 = 190 \\ 
2(0.96)(0.04) = 0.0768, & \ell = 95 - 5 = 90 \\ 
(0.96)^2 = 0.9216, & \ell = -5 - 5 = -10 
\end{cases}$$

with MGF

$$M_{L_1 + L_2}(z) = \mathbb{E}[e^{z(L_1 + L_2)}] = 0.0016 e^{190z} + 0.0768 e^{90z} + 0.9216 e^{-10z}$$

Running this minimization w.r.t. $z$ in R:

```{r}
fn=function(z){log((0.0016*exp(190*z)+0.0768*exp(90*z)+0.9216*exp(-10*z))/0.05)/z}
optimise(fn, c(0,1) )
```

we get that the minimum is $EVaR_{0.05}(L_1 + L_2) = 122.0294$,
occurring at $z = 0.03841828$.

Note that this EVaR is subadditive, since

$$\text{EVaR}_{0.05}(L_1 + L_2) \leq \text{EVaR}_{0.05}(L_1) + \text{EVaR}_{0.05}(L_2)$$

$$\Leftrightarrow 122.0294 \leq 2 \times 92.10402 = 184.208$$
