---
title: "Statistical Arbitrage"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Pairs Trading and Cointegration Analysis

## 1.1.Get Historical Stock Data

```{r}
library(tseries)
library(zoo)

start.date=as.Date("2000-01-01"); end.date=as.Date("2008-12-31")
P1=get.hist.quote(instrument = "XOM", start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T)
P2=get.hist.quote(instrument = "CVX", start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T)
P=merge(P1,P2, all=FALSE); P1=P[,1]; P2=P[,2]
lP1=log(P1); lP2=log(P2)
logRat= lP1-lP2 # log-ratio of prices
acf(coredata(logRat), lwd=2) # log-ratio autocorrelation function
```

-   `acf(coredata(logRat), lwd=2)`: Plots autocorrelation of log price
    ratio to understand temporal dependencies.

    -   `coredata(logRat)`: extracts just the numerical data (the actual
        values), ignoring the time/index.

        -   Because `acf()` works on regular numeric vectors, not all
            time series classes like `xts`. If `logRat` is an `xts`
            object, you'd need to pass its core data (the numeric
            values) to `acf()`—hence `coredata(logRat)`.

    -   `acf(...)`: computes and plots the autocorrelation function.

## 1.2.Plot Log Prices and Spread

```{r}
library(tseries);library(zoo)

#####
start.date=as.Date("2000-01-01"); end.date=as.Date("2008-12-31")

P1=get.hist.quote(instrument = "XOM", start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T)
P2=get.hist.quote(instrument = "CVX", start = start.date, end=end.date, quote = "AdjClose", retclass = "zoo", quiet = T)
P=merge(P1,P2, all=FALSE); P1=P[,1]; P2=P[,2]
lP1=log(P1); lP2=log(P2)

#####
plot( lP1, lwd=2); # Plot log-prices from series lP1
lines( lP2 + mean(lP1-lP2), lwd=2, col=2) # Add adjusted lP2 to the plot

#### Draws ±2 standard deviation bands around the mean.
logRat= lP1-lP2 # log-ratio of prices
plot( logRat, lwd=2); 
abline(h= mean(logRat), lwd=2, col=4)
abline(h=c(-2,2)*sd(logRat) + mean(logRat), col=4, lwd=2, lty=2) 

```

## 1.3.Pairs Trading Logic

```{r}

### 1.1 sets thresholds for opening/closing trades.
open.lim=2
bail.lim=3.5

### 1.2.Standardizes log-ratio;
std.logRat=(logRat-mean(logRat))/sd(logRat)

### 2.Initializes trading status and profit tracking.
trade.ind=0 # trade indicator: 0 means no position, 1 means buy 1st & sell 2nd stock, -1 means sell 1st & buy 2nd stock
n=length(logRat)
profit=rep(0,n)
t.open=t.close=NULL

### 3.Main Trading Loop
for(i in 1:n){
  # Close position if mean-reversion occurred or exceed bail limit
  if( (trade.ind*std.logRat[i]<0) | (abs(std.logRat[i])>bail.lim) ){
    trade.ind=0  # No open trade
    profit[i]=profit[i]+n1*P1[i]+n2*P2[i] # Realize profit/loss
    t.close=c(t.close,i)  # Log close time
  }
  
  # Open position if no current trade & log-ratio deviates exceed threshold
  if( trade.ind==0 & abs(std.logRat[i])>open.lim ){
    trade.ind=sign(coredata(std.logRat[i])) # Direction of trade (+1 or -1)
    n1=-trade.ind/coredata(P1[i]) # Short or long asset 1
    n2=trade.ind/coredata(P2[i]) # Long or short asset 2
    t.open=c(t.open,i) # Log open time
  }
}

### 4.plots cumulative profit over time
profit=zoo(profit,index(P1))
plot(cumsum(profit), lwd=2)
```

```{r}

### 5.Visualizes Spread Chart with Trade Signals

plot( logRat, lwd=2);  # log-price ratio between 2 assets (spread) as time series
abline(h= mean(logRat), lwd=2, col=4) # long-run average (blue).
abline(h=c(-2,2)*sd(logRat) + mean(logRat), col=4, lwd=2, lty=2) # standard deviation bands (dashed blue).
abline(v=index(P1[t.open]), lwd=2, col=3) # Entry points (Green lines)
abline(v=index(P1[t.close]), lwd=2, col=2) # Exit points (Red lines)

legend("topright", legend=c("Mean", "±2 SD", "Entry", "Exit"),
       col=c(4, 4, 3, 2), lty=c(1, 2, 1, 1), lwd=2)
```

## 1.4.Cointegration Tests

### 1.4.1.Two-step Engle-Granger method

-   Step 1: Find the potential long-run relationship

-   Step 2: Test if the difference from that relationship (residuals) is
    stationary.

    -   Phillips–Perron test : $H_0$: The series has a unit root (i.e.,
        it's non-stationary) ; $H_1$: The series is stationary

-   If Yes =\> the two series are cointegrated=\> usable for strategies
    like pairs trading.

```{r}

#### 1.regress P2 on P1
out=lm(lP2~lP1) 
plot(lP1,lP2); abline(out, lwd=2, col=2)
plot(out$res, lwd=2)

PP.test(coredata(out$res)) # Testing Residuals for Stationarity (Phillips-Perron Test)


#### 2.regres P1 on P2 & test residuals for Stationarity
PP.test(coredata(lm(lP1~lP2)$res))
```

-   `out = lm(lP2 ~ lP1)`: This runs a **linear regression** of the
    log-price of CVX (`lP2`) on the log-price of XOM (`lP1`): The fitted
    equation is:
    $$  \text{lP2} = \beta_0 + \beta_1 \cdot \text{lP1} + \epsilon  $$
-   where $\epsilon$ represents the residuals (the differences between
    the observed values and those predicted by the regression). If lP1
    and lP2 are **cointegrated**, the residuals $\epsilon$ should be
    **stationary** (i.e., their statistical properties do not change
    over time). Stationary residuals indicate that any deviation from
    the equilibrium is temporary and mean-reverting.

### 1.4.2.Phillips-Ouliaris Cointegration Test Method

```{r}


po.test(cbind(lP1,lP2)) # Phillips-Ouliaris test

```

# 2.Simulated Time Series and Stationarity Analysis

## 2.1.Simulated Time Series and Autocorrelations

AR(1) stands for Autoregressive model of order 1. It's a time series
model where the current value depends on its immediately previous value:

$$X_t = a \cdot X_{t-1} + \varepsilon_t$$

-   $a$ is the autoregressive coefficient (also often denoted by
    $\phi$); $\varepsilon_t$ is white noise (random shocks at time $t$)
-   If $a = 1$, it's a random walk (non-stationary).
-   If $|a| < 1$, the process is stationary, meaning it doesn't drift
    off to infinity over time and has a constant mean and variance.

```{r}
# Autocorrelations for different models

par(mfrow=c(3,2), mar=c(2,2,2,2))


### 1.Random walk (non-stationary)
X1=cumsum(rnorm(1000)) 
plot(X1, type='l', lwd=2, xlab="", ylab="")
acf(X1, lwd=2, main="")

### 2.stationary AR(1) with a=.95
X2=arima.sim(n=1000,list(ar=.95))
plot(X2, lwd=2)
acf(X2, lwd=2, main="")

### 3.stationary AR(1) with a=-.95
X3=arima.sim(n=1000,list(ar=-.95)) 
plot(X3, lwd=2)
acf(X3, lwd=2, main="")

par(mfrow=c(1,1))

```

For example an AR(1) model with coefficient $a = 0.95$, i.e.:

$$X_t = 0.95 \cdot X_{t-1} + \varepsilon_t$$

-   Since the value of $a$ is close to 1, the process exhibits strong
    persistence, meaning it remembers past values quite a bit.

## 2.2.Stationarity Tests (Phillips-Perron)

Phillips–Perron test on the time series.

$H_0$: The series has a unit root (i.e., it's non-stationary) ; $H_1$:
The series is stationary

```{r}
PP.test(X1)
PP.test(X2)
PP.test(X3)
```

p-value \> 0.05 =\> fail to reject the null =\> `X1` is likely
non-stationary.

p-value \< 0.05 =\> reject the null =\> `X2`, `X3` is stationary.

# 3.Two ETF Pairs Trading with Training Period Std Dev

Implement a pairs trading strategy of your own, using daily stock data
from Yahoo Finance. More specifically, find two ETF’s that are highly
correlated (e.g. both track the same index or industry) and download 10
years of historical data. Use the first 5 years to select the parameters
of your strategy, and the last 5 years to test it. Plot the
log-difference of the series over the entire period, as well as the
cumulative profit of the strategy.

## 3.1.ETF's data and Log-Ratio Plotting

We are trading “SPY” against “IVV”; both are large ETFs tracking the
S&P500 index. The plot of the log-ratio of their prices is shown below:

```{r}
library(zoo); library(tseries)

ETF.names=c('SPY','IVV')

N.ETF=length(ETF.names)
S=list() # S stores the time series for each ETF.
for(i in 1:N.ETF){
  print(paste(i, ' - ', ETF.names[i]))
  S[[i]]=get.hist.quote(ETF.names[i], start='2006-01-03', end='2016-12-30', quote='AdjClose')
}

names(S)=ETF.names

# Split data in half: first half for training, second half for testing.
n=length(S[[1]]); n.half=n%/%2 

# Plot log-ratio of prices: log(S1/S2).
plot(log(S[[1]]/S[[2]]), type='l'); abline(v=index(S[[1]])[n.half], lty=2, lwd=2)
```

Note that the log-ratio exhibits highly mean reverting behavior in the
training period (left half), especially around the Fall 2008 financial
crisis. In the testing period (right half), there are less intense
fluctuations, and an adjustment (jump) at the beginning of 2016.

## 3.2.Backtesting Pairs Trading Using Standardized Log-Price Ratio

Use the first 5 years to select the parameters of your strategy, and the
last 5 years to test it. Plot the log-difference of the series over the
entire period, as well as the cumulative profit of the strategy.

The pairs trading strategy employed involves opening a position at 2
standard deviations, closing a position at the mean, and bailing-out a
position at 4 standard deviations (where the **standard deviation is
calculated based on the first half of data**). The resulting strategy
and cumulative gain (per \$1 position in either stock) are shown below:

```{r}

## INPUT
open.lim=2 
bail.lim=4


### 1.Train Test Split -> std log-price ratio

S.1.train=coredata(S[[1]][1:n.half])
S.2.train=coredata(S[[2]][1:n.half])

S.1.test=coredata(S[[1]][(n.half+1):n])
S.2.test=coredata(S[[2]][(n.half+1):n])

profit.PP=rep(0,length(S.1.test))

logRatio.train=log(S.1.train/S.2.train)
logRatio.test=log(S.1.test/S.2.test)

std.logRatio.test=(logRatio.test-mean(logRatio.train))/sd(logRatio.train)

### 2.Pairs Trading
trade.ind=0 # trade indicator: 0 means no position, -1 means buy 1st & sell 2nd stock, 1 means sell 1st & buy 2nd stock
bail.ind=0 # bail-out indicator

t.open=t.close=NULL
for(j in 1:length(logRatio.test)){
  # check for closing an open position
  if( (trade.ind*std.logRatio.test[j]<0 | abs(std.logRatio.test[j])>bail.lim) & bail.ind==0 ){
    trade.ind=0
    profit.PP[j]=profit.PP[j]+n1*S.1.test[j]+n2*S.2.test[j]
    t.close=c(t.close,j)
    if( abs(std.logRatio.test[j])>bail.lim ){
      bail.ind=1 # bail out position, & stop trading
    }
  }
  # check for opening a position
  if( (trade.ind==0) & (abs(std.logRatio.test[j])>open.lim) & (bail.ind==0) ){
    trade.ind=sign(std.logRatio.test[j])
    n1=-trade.ind/S.1.test[j]
    n2=trade.ind/S.2.test[j]
    t.open=c(t.open,j)
  }
}


### 3.Plotting Results
par(mfrow=c(1,2))
plot(cumsum(profit.PP), type='l', ylab="cumulative profit per $1 invested", lwd=2)

plot(std.logRatio.test, type='l', lwd=2, ylim=c(-4,4))
abline(h=c(-2,2), col=5); abline(h=c(-4,0,4), col=4)
abline(v=t.open, col=3, lty=2); abline(v=t.close, col=2, lty=2)

```
