---
title: "Monte Carlo Simulation"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    self_contained: true
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

This is sourced from STAD70 course practice questions and sample R code
taught by professor Sotos. If you have any questions/concerns/comments
feel free to email me:
[cristal.wang111\@gmail.com](mailto:cristal.wang111@gmail.com){.email}.

# 1.Brownian Motion 1D (GBM)

Consider the arithmetic Brownian motion $X_t = \mu t + \sigma W_t$,
where $\{W_t\}$ is standard Brownian motion. Find the conditional
distribution of\
$$
\begin{bmatrix} 
X_{t_1} \\ 
\vdots \\ 
X_{t_n} 
\end{bmatrix} | X_1 = x, \quad \text{where } 0 < t_1 < \ldots < t_n < 1,
$$ also known as the multivariate Brownian bridge. Write the parameters
of the distribution in terms of $\mu, \sigma, t_1, \ldots, t_n$.

**Solution:** The (unconditional) joint distribution is\
$$
\begin{bmatrix} 
X_{t_1} \\ 
\vdots \\ 
X_{t_n} \\
X_1
\end{bmatrix} \sim N \left( \mu \begin{bmatrix} 
t_1 \\ 
\vdots \\ 
t_n \\
1
\end{bmatrix}, \quad \sigma^2 \begin{bmatrix} 
t_1 & \cdots & t_1 & t_1 \\ 
\vdots & \ddots & \vdots & \vdots \\ 
t_1 & \cdots & t_n & t_n \\ 
t_1 & \cdots & t_n & 1 
\end{bmatrix} \right)
$$

Using the Normal conditional distribution formula, we get:\
$$
\begin{bmatrix} 
X_{t_1} \\ 
\vdots \\ 
X_{t_n} 
\end{bmatrix} | (X_1 = x) \sim N \left( \mu \begin{bmatrix} 
t_1 \\ 
\vdots \\ 
t_n 
\end{bmatrix} + \begin{bmatrix} 
t_1 \\ 
\vdots \\ 
t_n 
\end{bmatrix} (x - \mu), \quad \sigma^2 \left( \begin{bmatrix} 
t_1 & \cdots & t_1 \\ 
\vdots & \ddots & \vdots \\ 
t_1 & \cdots & t_n 
\end{bmatrix} - \begin{bmatrix} 
t_1 \\ 
\vdots \\ 
t_n 
\end{bmatrix} \begin{bmatrix} 
t_1 & \cdots & t_n 
\end{bmatrix} \right) \right)
$$

$$
\sim N \left( x \begin{bmatrix} 
t_1 \\ 
t_2 \\ 
\vdots \\ 
t_n 
\end{bmatrix}, \quad \sigma^2 \begin{bmatrix} 
t_1(1-t_1) & t_1(1-t_2) & \cdots & t_1(1-t_n) \\ 
t_1(1-t_2) & t_2(1-t_2) & \cdots & t_2(1-t_n) \\ 
\vdots & \vdots & \ddots & \vdots \\ 
t_1(1-t_n) & t_2(1-t_n) & \cdots & t_n(1-t_n) 
\end{bmatrix} \right)
$$

## 1.1.Unconditional BM (ABM,GBM)

Consider the geometric Brownian motion
$S_t = 50 \exp \{ .03t + .2W_t\}$; fix $n = 100$ and let
$t_i = i/(n+1), \quad i = 0, \ldots, n$. Write R code that simulates and
plots a conditional path of the process, sampled at times
$\{t_i\}_{i=0}^{n+1}$, given that $W_1 = 1$. Use the result of the
previous question combined with the Cholesky decomposition for
generating the correlated values of $W$ at $(t_1, \ldots, t_n)$.

```{r}

### unconditional ABM & GBM (Not given W1=1)

set.seed(1234567890)
S0=50
mu=.03 # std BM: mu=0
sigma=.2 # std BM: sigma=1
n=100 # number of interior points
t.i=(1:n+1)/(n+1) # Time points from (1/101, 2/101, ..., 100/101,1)
 
MU.X=t.i*mu # mean 
mat1=matrix(t.i,n,n)
mat2=matrix(t.i,n,n,byrow=T)
SIGMA.X= (pmin(mat1,mat2))*sigma^2 # ABM covariance matrix
L=chol(SIGMA.X) # Cholesky decomposition of SIGMA.X

Z=rnorm(n)
X=MU.X + t(L)%*%Z #ABM
S=S0*exp(X) #GBM
plot(c(S0,S), type='l') # Plot GBM

```

## 1.2.Brownian Bridge (ABM, GBM)

```{r}

### BM (given W1=1)

set.seed(1234567890)
S0=50
mu=.03
sigma=.2
n=100 # number of interior points
t.i=(1:n)/(n+1) # Time points from (1/101, 2/101, ..., 100/101), exclude 0 and 1
 
X1=mu+sigma*1 # Xt=u*t+sigma*Wt
S1=S0*exp(X1)  # Final asset price at time 1
MU.X=t.i*X1 # conditional mean 
mat1=matrix(t.i,n,n)
mat2=matrix(t.i,n,n,byrow=T)
SIGMA.X= (pmin(mat1,mat2) - t.i %x% t(t.i))*sigma^2 # conditional covariance matrix
L=chol(SIGMA.X) # Cholesky decomposition of SIGMA.X

Z=rnorm(n)
X=MU.X + t(L)%*%Z #ABM
S=S0*exp(X) #GBM
plot(c(S0,S,S1), type='l')
```

```{r}

### BM (given Wb=a)
```

# 2.Correlated Brownian Motion

## 2.1.Simulating Correlated Standard BM

Simulating and Visualizing standard Brownian Motion paths with positive
and negative correlations

```{r}
##### 
# Correlated Standard BM: u=0, sigma=1

m=250 # number of steps

Z1=rnorm(m) 
W1=c(0, cumsum(Z1*sqrt(1/m))) # 1st BM path
plot((0:m)/m,W1,type='l', lwd=2, ylim=c(-1.5,1.5)) # 1

rho=+.8 
Z2=rho*Z1+sqrt(1-rho^2)*rnorm(m)
W2=c(0,cumsum(Z2*sqrt(1/m))) # 2nd BM path w/ rho=.8
lines((0:m)/m,W2, lwd=2, col=2)

rho=-.8 
Z2=rho*Z1+sqrt(1-rho^2)*rnorm(m)
W2=c(0,cumsum(Z2*sqrt(1/m))) # 2nd BM path w/ rho=-.8
lines((0:m)/m,W2, lwd=2, col=3)


legend("topright",legend=c("W1 (Standard BM)", "rho = +0.8", "rho = -0.8"),col=c(1,2,3),lty=1,lwd=2,bty="n")
```

## 2.2.Simulate Multivariate Geometric BM

**Simulating Correlated 3D Geometric BM**

```{r}
##### 
# Correlated 3D Geometric BM

### 1.INPUT - parameters

m = 250             # number of steps
Dt = 1 / m          # step size
Rho = matrix(0.4, 3, 3); diag(Rho) = 1  # correlation matrix
V = c(0.2, 0.3, 0.25)                   # volatilities (sigma, std dev)
Sig = Rho * (V %*% t(V))                # covariance matrix
L = chol(Sig)                           # Cholesky factorization
r = 0.02                                # risk free rate
Drft = r - V^2 / 2                      # drift term
S0 = 75                                 # Assume all assets start at 75

### 2.Simulates paths for all assets bt correlated Brownian motion.

Z=matrix( rnorm(m*3) , m, 3)
S=S0*exp(apply(Drft*Dt+ Z%*%L*sqrt(Dt),2, cumsum)) # Generate 3D path


### 3.Plots all GBM asset paths on the same time scale.

plot((0:m)/m, c(S0,S[,1]), type='l', lwd=2, ylim=c(60,105)) # Asset 1
lines((0:m)/m, c(S0,S[,2]), lwd=2, col=2) # Asset 2
lines((0:m)/m, c(S0,S[,3]), lwd=2, col=3) # Asset 3
legend("topright", legend = c("Asset 1", "Asset 2", "Asset 3"),  col = c(1, 2, 3),  lwd = 2, bty = "n")  # bty="n" removes the box
```

-   The system $\{S(t)\}$ follows the SDE: $$
    dS(t) = \mu \circ S(t) \, dt + \sigma \circ S(t) \, d\mathbf{W}(t)
    $$

    The solution is given by $S(t) = \exp\{X(t)\}$, where: $$
    dX(t) = \left( \mu - \frac{\sigma^2}{2} \right) dt + \sigma \, d\mathbf{W}(t)
    $$

    Generate geometric BM variates as: $$
    S(t_i) = S(t_{i-1}) \circ \exp\left\{ \left( \mu - \frac{\sigma^2}{2} \right) \Delta t + L Z_i \sqrt{\Delta t} \right\}, \quad i = 1, \ldots, m
    $$

    -   Where: $Z_i \stackrel{iid}{\sim} N_d(0, I)$ (standard
        multivariate normal) ;
        $LL^T = \Sigma = (\sigma \sigma^T) \circ \rho$ (Cholesky
        decomposition of covariance matrix) ; $\circ$ denotes
        element-wise product - $\rho$ is the correlation matrix

    -   $\mu$: Vector of drift rates

    -   $\sigma$: Vector of volatilities

    -   $\mathbf{W}(t)$: Vector of correlated Brownian motions

    -   $\Delta t$: Time step size

    -   $L$: Lower triangular Cholesky factor

# 3.Black-Scholes Pricing Function

## 3.1.European Options

```{r}
#####
# Function to compute Black-Scholes prices 
# of European call & put options

BSprice=function(S, X, r, M, v){
  # Returns 2-column matrix: Col1: Call prices; Col2: Put prices
  # arguments: Asset price (S0), Strike (X), risk-free rate (r)
  # Maturity (M), volatility (v)
  
  d1=(log(S/X)+(r+0.5*v^2)*M)/(v*sqrt(M))
  d2=d1-v*sqrt(M)
  
  call.pr= S*pnorm(d1) - X*exp(-r*M)*pnorm(d2)
  put.pr=X*exp(-r*M) * pnorm(-d2) - S*pnorm(-d1)
  
  return(cbind(call.pr,put.pr)) 
}


prices = BSprice(50:100,75,.02,.5,.2)

plot(50:100, prices[,1], type='l', xlab="S_0", ylab="C_0",
     main="European Call vs Asset Price ")
plot(50:100, prices[,2], type='l', xlab="S_0", ylab="P_0",
     main="European Put vs Asset Price ")

```

call: payoff=$(S_T-K)_+=S_T\cdot I_{(S_T\geq K)}-KI_{(S_T\geq K)}$,
price=$S_0N(d_1)-Ke^{-rT}N(d_2)$

put: payoff=$(K-S_T)_+$, price= $Ke^{-rT}N(-d_2)-S_0N(-d_1)$

## 3.2.General Case

AoN1(asset_or_nothing): Payoff= $S_T\cdot I_{(S_T\geq K)}$,
Price=$S_0N(d_1)$

CoN1(cash-one-dollar_or_nothing): Payoff= $I_{(S_T\geq K)}$,
Price=$e^{-rT}N(d_2)$

--

AoN2(other direction, asset_or_nothing): Payoff=
$S_T\cdot I_{(S_T \leq K)}$,Price=$S_0N(-d_1)$

CoN2(other direction, cash-one-dollar_or_nothing): Payoff=
$I_{(S_T\leq K)}$, Price=$e^{-rT}N(-d_2)$

```{r}


General_BSprice=function(S, X, r, M, v){
  # Returns 2-column matrix: Col1: Call prices; Col2: Put prices
  # arguments: Asset price (S0), Strike (X), risk-free rate (r)
  # Maturity (M), volatility (v)
  
  d1=(log(S/X)+(r+0.5*v^2)*M)/(v*sqrt(M))
  d2=d1-v*sqrt(M)
  
  AoN1 = S*pnorm(d1); CoN1 = exp(-r*M)*pnorm(d2)
  AoN2 = S*pnorm(-d1); CoN2 = exp(-r*M)*pnorm(-d2)
  
  # call.pr= AoN1-X*CoN1 = S*pnorm(d1) - X*exp(-r*M)*pnorm(d2)
  # put.pr= X*CoN2-AoN2 = X*exp(-r*M)*pnorm(-d2) - S*pnorm(-d1)
  
  return(cbind(AoN1,CoN1,AoN2,CoN2)) 
}



prices = General_BSprice(50:100,75,.02,.5,.2)

plot(50:100, prices[,1], type='l', xlab="S_0", ylab="AoN1",
     main="Asset_or_Nothing vs Asset Price ")
plot(50:100, prices[,2], type='l', xlab="S_0", ylab="CoN1",
     main="Cash_or_Nothing vs Asset Price ")


```

**Based on Interval**

InSa_b(interval asset): payoff=$S_T\cdot I_{(a\leq S_T\leq b)}$,
price=$S_0(N(d^a_1)-N(d^b_1))$

InCa_b(interval one-dollar-cash): payoff=$I_{(a\leq S_T\leq b)}$,
price=$e^{-rT}(N(d^a_2)-N(d^b_2))$

```{r}
a_b_BS_Price = function(a, b, S0, r, M, v){
  InSa_b = General_BSprice(S0,a,r,M,v)[,1]-General_BSprice(S0,b,r,M,v)[,1]
  InCa_b = General_BSprice(S0,a,r,M,v)[,2]-General_BSprice(S0,b,r,M,v)[,2]
  return(cbind(InSa_b,InCa_b))
}


prices = a_b_BS_Price(80,85,10:100,.02,.5,.2)


plot(10:100, prices[,1], type='l', xlab="S_0", ylab="AoN1",
     main="InSa_b vs Asset Price ")

```

# 4.European Call - Monte Carlo Simulation for Option Pricing

```{r}
#####
# Alternatively, you can use the function EuropeanOption() from the RQuantLib library
# install.packages("RQuantLib")
# library(RQuantLib)

set.seed(1234567890)

#####
# European option pricing by simulation 

### 1.INPUT - Parameters
S0=70 # asset price at time 0
v=.2 # annual volatility
r=.02 # annual risk-free rate
K=75 # strike price
M=.5 # Time to maturity (in years)
n=100000 # number of variates

### 2.Simulate Asset Price at Maturity
Z=rnorm(n) # Standard Normal variates
S=S0*exp( (r-v^2/2)*M + v*sqrt(M)*Z ) # asset price variates @ T

### 3.Histogram of Simulated Asset Prices
hist(S, main="", prob=TRUE) # asset price variate histogram
lines( seq(30,130,.02), dlnorm(seq(30,130,.02)/S0, # theoretical logNormal density
                               (r-v^2/2)*M, v*sqrt(M) )/S0, col=2, lty=2, lwd=2)




```

-   `S=S0exp((r-v^2/2)*M + v*sqrt(M)*Z)`: This simulates the asset price
    `S` at maturity using the geometric Brownian motion (GBM)
    formula:$S_{T} = S_{0} \cdot \exp\left( \left( r - \frac{1}{2} v^{2} \right) T + v \sqrt{T} Z \right)$

    Where: - $S_{T}$: Stock price at time $T$, - $S_{0}$: Initial stock
    price, - $r$: Risk-free interest rate, - $v$: Volatility, - $T$:
    Time to maturity, - $Z$: Standard normal random variable.

-   `dlnorm(seq(30,130,.02)/S0,(r-v^2/2)*M, v*sqrt(M) )/S0` :

    -   The log-return of the stock price follows a normal distribution:

        $$
        \begin{align*}
        \log(S_T/S_0) \sim \mathcal{N} \left( (r - \frac{1}{2} v^2) T, v \sqrt{T} \right)
        \end{align*}
        $$

        Where: - $S_{T}$: Stock price at time $T$, - $S_{0}$: Initial
        stock price, - $r$: Risk-free interest rate, - $v$:
        Volatility, - $T$: Time to maturity, - $Z$: Standard normal
        random variable.

    -   `dlnorm(x, meanlog, sdlog)`: log-normal probability density
        function.

```{r}

### 4.lognormal QQ-Plot to Check Distribution Fit
qqplot( S/S0, qlnorm(ppoints(n), (r-v^2/2)*M, v*sqrt(M) ) )
abline(0,1, lwd=2, col=2)

```

```{r}
### 5.Mean and Standard Deviation Check

mean(S)  # must be approx. S0*exp(r*M)
S0*exp(r*M)

sd(S) # must be approx. S0*sqrt( exp(2*r*M) * (exp(v^2*M)-1) )
S0*sqrt( exp(2*r*M) * (exp(v^2*M)-1) )
```

```{r}

### 6.1.Monte Carlo Estimate of Call Option Price
call.MCprice=mean( exp(-r*M)*pmax(0,S-K) ) # Monte-Carlo (MC) call price

### 6.2.Compare to Exact Black-Scholes Price
call.BSprice=BSprice(S0,K,r,M,v)[1] # Exact Black-Scholes price
# call.BSprice=EuropeanOption("call",S0,75,0,.02,.5,.2)$value  # using RQuantLib

### 7.MC Price Convergence Plot
MC.prices=cumsum( exp(-r*M)*pmax(0,S-K) ) / (1:n) # average of Monte Carlo estimate of the call option price.
S_C=sd(exp(-r*M)*pmax(0,S-K)) # Standard deviation of all discounted payoffs.

a=0.95;Za = qnorm(1-(1-a)/2) ;ind=seq(50,n,by=50)
plot( ind, MC.prices[ind], ylim=c(1.25,3.75),lwd=2, type='l') # Monte Carlo price
abline(h=call.BSprice[1], col=2, lwd=2) # True B-S price
lines(ind, call.BSprice[1]+ Za*S_C*sqrt(1/ind), lwd=2, lty=2, col=2) # Upper 95% CI
lines(ind, call.BSprice[1]- Za*S_C*sqrt(1/ind), lwd=2, lty=2, col=2) # Lower 95% CI
legend( 5000,3.7, c("MC price", "True (B-S) price", "95% C.I."), lty=c(1,1,2), col=c(1,2,2), lwd=c(2,2,2) ) 
```

-   `call.MCprice=mean( exp(-r*M)*pmax(0,S-K) ):`

    -   The option pricing formula is given by:
        $$ C_{0} = e^{-rT} \cdot \mathbb{E}[\max(S_{T} - K, 0)] $$

-   `2*S_C*sqrt(1/ind)`:

    -   Approximates a 95% confidence interval using the 2-sigma rule
        (since $Z_{0.975} \approx 1.96$, people often just use 2).

# 5.Exchange Option - Monte Carlo Pricing

**Exchange option**

Suppose $S_1(t)$ and $S_2(t)$ are the prices of two risky assets at time
$t$, each with constant continuous dividend yield $q_i$. The exchange
option gives the right (but not obligation) to exchange asset 2 for
asset 1 at maturity $T$, with payoff:
$$ C(T) = \max(0, S_1(T) - S_2(T)) $$

**Combined Volatility**

If the volatilities are $\sigma_i$ and $\rho$ is their correlation
coefficient:
$$ \sigma = \sqrt{\sigma_1^2 + \sigma_2^2 - 2\sigma_1\sigma_2\rho} $$

Margrabe's formula states that the fair price for the option fair price
at time 0 is: $$ C(0) = e^{-q_1T}S_1(0)N(d_1) - e^{-q_2T}S_2(0)N(d_2) $$

Where: $N(\cdot)$ = Standard normal CDF \| $q_1, q_2$ = Dividend yields
of $S_1$ and $S_2$

$d_1 = \frac{\ln(S_1(0)/S_2(0)) + (q_2 - q_1 + \sigma^2/2)T}{\sigma\sqrt{T}}$ -
$d_2 = d_1 - \sigma\sqrt{T}$

```{r,warning=FALSE}
##### 
# MC Pricing of Exchange Option

### 1.INPUT - parameters
n=100000 # number of paths (no steps needed)
TT=1 # maturity
rf=.03 # risk free rate
sigma1=0.4; sigma2=0.3 # volatilities
V=c(sigma1, sigma2) 
rho=.3 # correlation
Rho=matrix( c(1,rho,rho,1),2,2 )
Sig=Rho*(V%*%t(V)) # Covariance matrix (element-wise multiply)
L=chol(Sig) # Cholesky factorization of covariance matrix
Drft=rf-V^2/2 # risk-neutral drift for log-prices
S0=c(50,35) # Initial prices of Asset 1 and Asset 2

### 2.Simulate Terminal Prices
Z=matrix(rnorm(n*2), n, 2)
ST=exp( matrix(Drft*TT,n,2,byrow=TRUE)+ Z%*%L*sqrt(TT))*matrix(S0,n,2,byrow=TRUE)
payoff=pmax(ST[,1]-ST[,2],0) # Payoff of the exchange option

MC.price=cumsum(payoff*exp(-rf*TT))/(1:n) # Discounted average
ind=seq(100,n,100)
plot( ind, MC.price[ind], pch=16 )

# Use "Margrabe's formula" (https://en.wikipedia.org/wiki/Margrabe%27s_formula) 
# to find the exchange option's price based on the Black-Scholes formula
library(RQuantLib)
exact.price=EuropeanOption("call",S0[1],S0[2],0,0,TT, sqrt(sigma1^2 + sigma2^2 - 2*sigma1*sigma2*rho ) )$value  # using RQuantLib

abline(h=exact.price, lwd=2, col=2)

```

-   `EuropeanOption(type, underlying, strike, dividendYield, riskFreeRate, maturity, volatility)`:

    -   You set `riskFreeRate = 0` in this formula only because: You are
        pricing an **exchange option using Margrabe's formula**, which
        inherently assumes no discounting is needed — the two risky
        assets grow at the same rate under risk-neutral measure, so the
        risk-free rate cancels out in the math.

# 6.Binary Option - Monte Carlo & Exact Price

Consider a Binary option that pays off $S_T 1_{\{S_T \geq K\}} =
\begin{cases}
S_T, & S_T \geq K \\
0, & S_T < K
\end{cases}$ at maturity $T$. This is a European asset-or-nothing binary
call, and has a analytical price in the Black-Scholes model (i.e. for
constant interest rate and an underlying asset following Geometric
Brownian motion). Consider a model where the interest rate is $r = .05$
and the asset has initial price $S(0) = 100$ and volatility
$\sigma = .4$. Find the price of the asset-or-nothing call with strike
$K = 100$ and maturity $T = .5$ using Monte Carlo simulation with
$n = 10,000$. Report your point estimate and the approximate 95%
confidence interval for the price, and compare it to the exact price.

```{r}

set.seed(1234567890)
S0=100; r=0.05; sigma=.4; T= 0.5; K = 100

# exact price 

C.exact = General_BSprice(S0,K,r,T,sigma)[,1]
print( paste("Exact binary call price =", C.exact) )
```

```{r}

set.seed(1234567890)
S0=100; r=0.05; sigma=.4; T= 0.5; K = 100

# exact price 
d1 = ( log(S0/K) + ( r + sigma^2 / 2 ) * T ) / ( sigma * sqrt(T) )
C.exact = S0 * pnorm( d1 )
print( paste("Exact binary call price =", C.exact) )

# MC simulation 
n=10000
Z = rnorm( n )
ST = S0 * exp( ( r - sigma^2 / 2 ) * T + sigma * sqrt(T) * Z )
payoff_3 = exp( -r * T ) * ifelse( ST>K, ST, 0 )
mean_3 = mean(payoff_3); se_3 = sd(payoff_3) / sqrt(n)

a=0.95; CI_3 = mean_3 + c(-1,1) * qnorm(1-(1-a)/2) * se_3 
print( paste("MC binary call price =", mean_3) )
print( paste( c("MC binary call",a*100,'% CI =', CI_3), collapse = "  " ) )
```

-   `ifelse( ST>K, ST, 0 )`: This is equivalent to `(ST>=K)* pmax(0,ST)`

# 7.Rainbow Option (Multi-Assets) - Monte Carlo Pricing

A rainbow option is a derivative exposed to two or more underlying
assets. Consider a European rainbow option with payoff given by\
$$\left( \max_i \{ S_T^{(i)} \} - \frac{1}{d} \sum_{i=1}^d S_T^{(i)} \right),$$\
i.e. the payoff is the maximum final price minus the average final price
of all $d$ assets (note that the maximum/average is over assets, not
time).

Estimate the **price of this option using Monte Carlo simulation** with
$n = 10,000 \, \text{d-dimensional paths},$ and provide a 95% confidence
interval with your answer.

Assume that
$d = 5, K = 100, T = 1, r = 0.03, S^{(i)}(0) = 100, \, \forall i = 1, \ldots, d,$
and that the assets follow multivariate Geometric Brownian Motion:

$dS(t) = rS(t)dt + \sigma \circ S(t) \circ d\mathbf{W}(t)$ for
$\sigma = [.1\ .2\ .3\ .4\ .5]$ and Corr $(W_i(1), W_j(1)) = .3$,
$\forall i \neq j$

where $\circ$ is the Hadamard or element-wise matrix product.

(Note: there is no general analytical solution for rainbow options.)

```{r}

set.seed(1234567890)
n=10000; d=5; S0=100; M=1; r=.01 # - - |

Rho= matrix(.3,5,5); diag(Rho)=1 # correlation matrix - - |
V=c(1:5/10) # volatilities - - |
Sig=Rho*(V%*%t(V)) # covariance matrix - - |
L=chol(Sig) # Cholesky factorization of covariance matrix
Drft=matrix(r-V^2/2, n, d, byrow=TRUE) # drift
Z=matrix( rnorm(n*d) , n, d)
ST=S0*exp( Drft*M+ Z%*%L*sqrt(M) )


maxT=apply(ST,1,max)
avgT=apply(ST,1,mean)
payoff_4=exp(-r*M)*(maxT-avgT) # Payoff - - |

mean_4=mean(payoff_4)
se_4=sd(payoff_4)/sqrt(n)

a=0.95;CI_4=mean_4+c(-1,1)*qnorm(1-(1-a)/2)*se_4
print( paste("rainbow option price =", mean_4) )
print(  paste( c("rainbow option",a*100,'% CI =',CI_4), collapse = "  " ) )
```
